---
title: "ETV Cycle 3 — California County Populations from Wikipedia"
format: html
editor: visual
---

## Description

Wikipedia hosts a rich table of California counties and their populations on the [California counties Wikipedia page](https://en.wikipedia.org/wiki/List_of_counties_in_California). We will install `htmltab` from GitHub, use `htmltab` to extract data from the California Wikipedia table, clean it, and explore how county populations are distributed within California.

Each **row** represents one California county. Each **column** represents an attribute of that county.

------------------------------------------------------------------------

### Motivating Questions

**Subtask 1** — Which California counties have the largest populations?

**Subtask 2** — What does the distribution of California county populations look like?

**Subtask 3** — How are California county area and population related?

------------------------------------------------------------------------

## Learning Moment 0 — One-Time Package Installation

Run this chunk once if `htmltab` is not already installed on your machine. The package is no longer installed from CRAN in the usual way, so we use `remotes` to install it from GitHub.

```{r eval=FALSE}
install.packages("remotes")
remotes::install_github("htmltab/htmltab")
```

**Setup Q1.** What does `remotes::install_github()` do that `install.packages()` does not?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Learning Moment 1 — Libraries

```{r}
library(htmltab) ## Data Extraction  — scrape HTML tables with htmltab()
library(____)   ## Data Transformation — filter, mutate, arrange, slice_head
library(____)   ## Data Transformation — separate()
library(____)   ## Data Transformation — cleaning string columns
library(____)   ## Data Transformation — parse_number()
library(____)   ## Data Visualization
library(____)   ## Data Visualization — reordering factors
```

**Side Q1.** What is the difference between scraping a webpage and downloading a CSV file?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Extraction (E)

#### Learning Moment 2 — Scraping an HTML Table with `htmltab`

`htmltab()` reads a table directly from a webpage. The `which` argument tells R which table on the page to import.

```{r}
url <- "https://en.wikipedia.org/wiki/List_of_counties_in_California"

ca_df <- htmltab::htmltab(
  url,
  which = 2,
  rm_nodata_cols = FALSE
)

glimpse(ca_df)
```

```{r}
county_df <- ca_df
names(county_df)
```

**Q1.** What does the argument `which = 2` tell `htmltab()` to do?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

**Side Q2.** Why might a Wikipedia page have more than one HTML table?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Transformation (T)

#### Learning Moment 3 — Cleaning Column Names

```{r}
## Rename columns to clean lowercase names
county_df <- county_df |>
  rename(
    county     = ____,
    population = ____,
    area       = ____
  )


```

------------------------------------------------------------------------

#### Learning Moment 4 — Converting Population to Numeric

```{r}
county_df <- county_df |>
  mutate(population = ____(population))

glimpse(county_df)
```

**Q3.** Why do we use `parse_number()` rather than `as.numeric()`?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 5 — Splitting Area into Two Numeric Columns

The `area` column stores two measurements inside one text value:

```{r}
county_df |>
  select(county, area) |>
  slice_head(n = 2)
```

We want to turn values like `"738 sq mi(1,911 km)"` into two separate numeric columns: `area_sq_mi` and `area_sq_km`.

```{r}
county_df <- county_df |>
  separate(
    area,
    into = c("area_sq_mi_text", "area_sq_km_text"),
    sep = "____",
    remove = ____
  )

county_df <- county_df |>
  mutate(
    area_sq_mi = ____(area_sq_mi_text),
    area_sq_km = ____(area_sq_km_text)
  ) |>
  select(-area_sq_mi_text, -area_sq_km_text)

glimpse(county_df)
```

**Q4.** What does `separate()` do to the `area` column in this example?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

**Side Q4.** Why do we use `parse_number()` again after separating the area text?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Visualization (V)

#### Horizontal Bar Chart

```{r}
top_county_df <- county_df |>
  arrange(desc(____)) |>
  slice_head(n = ____)

ggplot(top_county_df,
       aes(x = ____, y = fct_reorder(____, ____))) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Top 15 California Counties by Population",
    x     = "Population",
    y     = "County"
  ) +
  theme_minimal()
```

**Q5.** What does `scales::comma` do inside `scale_x_continuous(labels = ...)`?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Histogram with a Log Scale

```{r}
ggplot(county_df |> filter(!is.na(____)),
       aes(x = ____)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  scale_x_log10(labels = scales::comma) +
  labs(
    title = "Distribution of California County Populations (Log10 Scale)",
    x     = "Population (log scale)",
    y     = "Number of Counties"
  ) +
  theme_minimal()
```

**Q6.** Why is a log scale useful for county population data?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

**Side Q5.** Why do we use `filter(!is.na(population))` before plotting?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

### Subtask 3 — Area and Population Scatter Plot

```{r}
ggplot(county_df |> filter(!is.na(____), !is.na(____)),
       aes(x = ____, y = ____)) +
  geom_point(color = "steelblue", alpha = 0.75, size = 2) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "California County Population by Area",
    x     = "Area (square miles)",
    y     = "Population (log scale)"
  ) +
  theme_minimal()
```

**Q7.** What does this scatter plot help us compare?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

### Subtask 4 — Student-Generated Question

Propose a question comparing California counties by population, square miles, square kilometers, or another variable in the Wikipedia table. Your instructor will select one for class exploration.

```{r}
## Your code here

```