---
title: "ETV Cycle 4 — Weather, Weather: API Requests, purrr, and Plot Types"
format: html
editor: visual
---

## Description

An **API** (Application Programming Interface) is a structured way for one program to request data from another program or web service. In this cycle, we will use the **Open-Meteo API** to request recent weather data, parse the JSON response, repeat the request for multiple cities with `purrr`, add date features with `lubridate`, and compare a few plot types.

------------------------------------------------------------------------

### Motivating Questions

**Subtask 1** — What do daily high and low temperature trends look like across several California cities?

**Subtask 2** — How does the relationship between daily temperature range and precipitation differ across cities?

**Subtask 3** — Are days with larger temperature ranges also days with more or less precipitation?

------------------------------------------------------------------------

## Setup — Libraries

```{r}
library(____)
library(____)
library(____)
library(____)
library(____)
library(____)
library(____)
```

::: callout-important
## The Extraction Is the Heavy Hitter

In this cycle, most of the intellectual work happens in **Extraction (E)**: understanding the API, building the request, checking the response, parsing JSON, and repeating the workflow for multiple cities.

**Reflection.** Why is the Extraction (E) step emphasized as the most important — the "heavy hitter" — part of this ETV cycle? Answer in 2–3 sentences.

> **Your answer:**
:::

------------------------------------------------------------------------

## Extraction (E)

#### Learning Moment 0 — What Is an API?

An **API** lets us ask a service for data in a structured way. Instead of manually downloading a spreadsheet, we send a request to a URL and include arguments that describe what data we want.

For this cycle, we use Open-Meteo:

<https://open-meteo.com/>

Open-Meteo is useful here because it does not require an API key and it can return daily weather variables for a specific latitude and longitude.

**Side Q1.** Why might an API be better than manually downloading a file?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 1 — Request, Response, and JSON

The basic API workflow has three pieces:

1.  A **request**: what we ask the API for.
2.  A **response**: what the API sends back.
3.  A **JSON body**: the text format that stores the returned data.

<!-- TODO: confirm image URL -->

![The API request -\> response -\> JSON workflow](api-json-workflow.svg){fig-alt="Diagram showing R code making an HTTP GET request to an API, receiving JSON, and converting it into an R data frame." width="90%"}

```{r}
# (a) making request
base_request <- ____(____)

# (b) by default GET
base_request
```

**Annotation (a).** What does `request(...)` do at this step?

> **Your answer:**

**Annotation (b).** Why is the default HTTP method GET here?

> **Your answer:**

**Q1.** What is JSON, and why is it common for API responses?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 2 — Planning the Query First

Before sending a request, write down the pieces the API needs. For weather data, we need a location, variables, units, time zone, and date range.

```{r}
# (a) location information
city_name <- ____
latitude <- ____
longitude <- ____

# (b) requested daily weather variables
daily_variables <- c(
  ____,
  ____,
  ____
)

city_name
daily_variables
```

**Annotation (a).** What does the `city_name` / `latitude` / `longitude` group define?

> **Your answer:**

**Annotation (b).** What does `daily_variables` define, and why store it separately?

> **Your answer:**

**Q2.** Why is it helpful to save query inputs as objects before building the request?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 3 — Doing One Query First

Now we build and send one request. This step is intentionally just one city, because it is easier to debug one working request before repeating it.

```{r}
single_city_request <- base_request |>
  req_url_query(
    # (a) from before
    latitude           = ____,
    longitude          = ____,
    daily              = ____,
    # (b) constant information
    temperature_unit   = ____,
    precipitation_unit = ____,
    timezone           = ____,
    past_days          = ____,
    forecast_days      = ____,
    .multi             = ____
  )

single_city_response <- single_city_request |>
  # (c)
  ____()

# (d)
resp_status(single_city_response)
```

**Annotation (a).** Which earlier objects are reused inside `req_url_query()`?

> **Your answer:**

**Annotation (b).** Why are `temperature_unit`, `precipitation_unit`, `timezone`, `past_days`, and `forecast_days` treated as constants here?

> **Your answer:**

**New Q1.** Explain the difference between `request()`, `req_url_query()`, and `req_perform()`.

| Function          | Your explanation |
|-------------------|------------------|
| `request()`       |                  |
| `req_url_query()` |                  |
| `req_perform()`   |                  |

**Q3.** What does an HTTP status code of `200` indicate?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 4 — Arguments Needed for `req_url_query()`

`req_url_query()` adds query parameters to the request URL. Each argument narrows or shapes the data returned by the API.

```{r}
# (a) JSON text -> R list
single_city_json <- single_city_response |>
  ____()

single_city_list <- single_city_json |>
  ____()

names(single_city_list)
```

```{r}
# (c)
single_city_daily <- single_city_list[[____]]

# (d) type and extraction: list -> data frame -> cleaned data frame
single_city_weather_df <- as.data.frame(single_city_daily) |>
  rename(date = ____) |>
  mutate(
    date = ____(date),
    city = ____,
    # (e)
    .before = 1
  )

glimpse(single_city_weather_df)
```

**Annotation (a).** Explain what `single_city_list` is and how it is structured.

> **Your answer:**

**Annotation (b).** What is the data type at each step, and what is being extracted?

> **Your answer:**

**Annotation (c).** Why do we extract `daily` specifically?

> **Your answer:**

**Annotation (e).** Why `.before = 1`? What does it control?

> **Your answer:**

**Side Q2.** What do the main `req_url_query()` arguments control?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

**Meta-reflection.** Why is Side Q2 asked here, and why is it placed at this point in the workflow?

> **Your answer:**

------------------------------------------------------------------------

#### Learning Moment 5 — A Small `map()` Moment, Then `pmap_dfr()`

**Say we want multiple cities — and how `map` helps us.**

Once we move from one city to several cities, we need a way to repeat the same operation without copying and pasting the same code block.

> **Your answer:** In your own words, why do we need a map function once we move from one city to many?

`purrr` map functions repeat the same operation over many inputs. Start small: `map_dbl()` below rounds every latitude.

```{r}
# (a)
cities <- data.frame(
  city      = c("San Luis Obispo", "San Francisco", "San Diego"),
  latitude  = c(35.2828, 37.7749, 32.7157),
  longitude = c(-120.6596, -122.4194, -117.1611)
)

```

**Do this blank (i).** Complete the `map_dbl()` call that rounds each latitude to one decimal place.

```{r}
cities |>
  mutate(latitude_rounded = ____)
```

<!-- Answer key: map_dbl(latitude, \(x) round(x, 1)) -->

When a function needs more than one input at the same time, use `pmap_dfr()`. It takes each row of a data frame, sends the row values into a function, and row-binds the data frames returned by that function.

```{r}
# (ii)
____(
  cities,
  # (iii)
  \(city, latitude, longitude) {
    data.frame(
      city = ____,
      location_text = paste0(____, ", ", ____)
    )
  }
)
```

**Q4.** What is the difference between `map_dbl()` and `pmap_dfr()` in the examples above?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 6 — Building the Final `get_weather()` Function

Now that one query works, we turn the steps into a function. The function takes one city, one latitude, and one longitude, then returns a cleaned daily weather data frame.

**Do it blank — Strategy.** The major pieces of `get_weather()` are scrambled below. Reorder them before reading the full function.

``` text
(a)
(b)
(c)
(d)
(e)
(f)
```

```{r}
get_weather <- function(city, latitude, longitude) {  # (a)
  # (b):
  url <- ""
  response <- request(____) |>
    req_url_query(
      latitude           = ____,
      longitude          = ____,
      daily              = c(____, ____, ____),
      temperature_unit   = ____,
      precipitation_unit = ____,
      timezone           = ____,
      past_days          = ____,
      forecast_days      = ____,
      .multi             = ____
    ) |>
    ____()

  # (c):
  ____(response)

  # (d):
  daily_data <- response |>
    ____() |>
    fromJSON() |>
    (\(x) x[[____]])()

  # (e):
  daily_weather_df <- as.data.frame(daily_data) |>
    rename(date = ____) |>
    mutate(
      date = ____(date),
      city = city,
      .before = 1
    )

    return(____)
}

# (f): 
weather_df <- ____(cities, get_weather)

glimpse(weather_df)
```

**Comment practice.** Add a one-line comment for each major block in the function: request build, `req_perform()`, JSON extraction, and data-frame cleaning.

``` text
# ___ 
# ___ 
# ___ 
# ___ 
# ___
```

------------------------------------------------------------------------

## Transformation (T)

#### Learning Moment 7 — Adding Date and Weather Features

```{r}
weather_df <- weather_df |>
  group_by(____) |>
  mutate(
    weekday         = ____(date, label = TRUE, abbr = TRUE) #,
    # week_start      = ____(date, unit = "week") #,
    # days_from_start = as.integer(date - ____(date)) #,
    # temp_range_f    = ____ - ____ ## We actual need this!
  ) |>
  ungroup()

# weather_df |>
#   select(____, ____, ____, ____, ____, ____) |>
#   slice_head(n = ____)
```

**Q6.** What do `wday()` and `floor_date()` help us do?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 8 — Pivoting Highs and Lows for Plotting

```{r}
weather_long_df <- weather_df |>
  pivot_longer(
    cols      = c(____, ____),
    names_to  = ____,
    values_to = ____
  ) # |>
  # mutate(
  #   temp_type = case_when(
  #     ____ == ____ ~ ____,
  #     ____ == ____ ~ ____
  #   )
  # )

weather_long_df |>
  select(____, ____, ____, ____) |>
  slice_head(n = ____)
```

**Q7.** What does `pivot_longer()` do here?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Visualization (V)




#### Learning Moment 9 — Line Plot: Temperature Over Time


**Subtask 1** — What do daily high and low temperature trends look like across several California cities?

```{r}
ggplot(____, aes(x = ____, y = ____, color = ____)) +
  geom_line(linewidth = ____) +
  facet_wrap(~ ____, ncol = ____) +
  scale_color_manual(values = c("Daily High" = ____, "Daily Low" = ____)) +
  labs(
    title = "Daily High and Low Temperatures by City",
    x     = "Date",
    y     = "Temperature (F)",
    color = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = ____, hjust = ____))
```

**Q8.** What kind of question is a line plot especially good at answering?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 10 — Scatter Plot: Temperature Range and Precipitation


**Subtask 2** — How does the relationship between daily temperature range and precipitation differ across cities?



```{r}
ggplot(____, aes(x = ____, y = ____, color = ____)) +
  geom_point(size = ____, alpha = ____) +
  labs(
    title = "Daily Temperature Range vs. Precipitation",
    x     = "Temperature Range (F)",
    y     = "Precipitation (inches)",
    color = "City"
  ) +
  theme_minimal()
```

**Q9.** What kind of question is a scatter plot especially good at answering?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 11 — Histogram: Distribution of Temperatures


**Subtask 3** — Are days with larger temperature ranges also days with more or less precipitation?


```{r}
ggplot(____, aes(x = ____, fill = ____)) +
  geom_histogram(bins = ____, alpha = ____, position = ____, color = ____) +
  facet_wrap(~ ____) +
  scale_fill_manual(values = c("Daily High" = ____, "Daily Low" = ____)) +
  labs(
    title = "Distribution of Daily Highs and Lows",
    x     = "Temperature (F)",
    y     = "Number of Days",
    fill  = NULL
  ) +
  theme_minimal()
```

**Side Q3.** What does a histogram show that a line plot does not show as directly?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

### Subtask 4 — Student-Generated Question

Change the cities, extend the weather variables, or choose a new plot type. Propose a weather-related question and visualize the answer.

```{r}
## Your code here

```