---
title: "widgets"
format: docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE,    # don't run code by default
  echo = TRUE,     # still show code unless you hide with echo=FALSE
  warning = FALSE, # suppress warnings
  message = FALSE, # suppress messages
  error = FALSE    # suppress errors
)
```

# Mini-Lesson 2: Shiny Widgets

## 0) what are widgets? (30s)

> “In Shiny, **widgets** are the controls your user interacts with: sliders, dropdowns, text boxes, checkboxes, file uploads, date pickers, etc. Every widget has two parts:
>
> 1.  A **UI function** like `sliderInput()` that goes in the UI.
> 2.  A **value** you can use in the server via `input$…`.”

------------------------------------------------------------------------


\newpage

## 1) simple demo app (90s)

Create `widgets_app.R`:

```{r}
library(shiny)

ui <- fluidPage(
  titlePanel("Widget Sampler"),

  # Sidebar for inputs
  sidebarLayout(
    sidebarPanel(
      sliderInput("num", "Number of points", 10, 500, 100),
      selectInput("dist", "Distribution",
                  choices = c("Normal" = "norm", "Uniform" = "unif",
                              "Exponential" = "exp")),
      checkboxInput("show_summary", "Show summary", TRUE),
      textInput("title", "Plot title", "My Plot"),
      dateInput("date", "Pick a date:", value = Sys.Date())
    ),

    # Main panel for outputs
    mainPanel(
      h4("Output"),
      plotOutput("hist"),
      verbatimTextOutput("summary"),
      textOutput("date_chosen")
    )
  )
)

server <- function(input, output, session) {
  data <- reactive({
    switch(input$dist,
           norm = rnorm(input$num),
           unif = runif(input$num),
           exp  = rexp(input$num))
  })


  
  output$hist <- renderPlot({
  ggplot(data.frame(x = data()), aes(x = x)) +
    geom_histogram(bins = input$bins, fill = "skyblue", color = "white") +
    labs(
      title = input$title,
      x = "Value",
      y = "Count"
    ) +
    theme_minimal(base_size = 14)
})

  output$summary <- renderPrint({
    req(input$show_summary)   # only show if box is checked
    summary(data())
  })

  output$date_chosen <- renderText({
    paste("You picked:", as.character(input$date))
  })
}

shinyApp(ui, server)
```

Run it:

```{r}
shiny::runApp("widgets_app.R")
```

------------------------------------------------------------------------
\newpage
## 2) widget categories (60s)

> “Shiny’s widgets fall into families:”

-   **Numeric**: `sliderInput`, `numericInput`.
-   **Text**: `textInput`, `passwordInput`, `textAreaInput`.
-   **Logical**: `checkboxInput`, `checkboxGroupInput`, `radioButtons`.
-   **Lists**: `selectInput`, `selectizeInput`.
-   **Dates**: `dateInput`, `dateRangeInput`.
-   **File / action**: `fileInput`, `actionButton`.
-   **Others**: `colourInput` (from `shinyWidgets` package), `downloadButton`, etc.

------------------------------------------------------------------------
\newpage
## 3) teaching tip (60s)

> “Every widget sets a value in the `input` list. For example:

-   `input$num` is the slider number,
-   `input$dist` is the dropdown choice,
-   `input$title` is whatever you typed in the text box.”

You can show students the full input object with:

```{r}
output$debug <- renderPrint(str(reactiveValuesToList(input)))
```

------------------------------------------------------------------------
\newpage
## 4) extension (60s)

> “Shiny has many extra widget libraries:”

-   **`shinyWidgets`**: pretty sliders, toggle switches, multi-select pickers.
-   **`bslib`**: bootstrap-themed inputs with modern styles.
-   **`shiny.semantic`**: widgets styled with Semantic UI.

Example (requires `shinyWidgets`):

```{r}
library(shinyWidgets)
pickerInput("var", "Choose variables", choices = names(mtcars),
            multiple = TRUE, selected = "mpg")
```

------------------------------------------------------------------------

\newpage

## 5) summary (30s)

> “Widgets are the *verbs* of your app—they let users tell your server what to do. Learn the common ones, know they map to `input$values`, and remember you can always add libraries to get more polished controls.”

------------------------------------------------------------------------

want me to design the **next lesson** as a “widget playground” (one app showing *all common widgets* side-by-side), so you or your students can interactively test each one?
