---
title: "outputs"
format: docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE,    # don't run code by default
  echo = TRUE,     # still show code unless you hide with echo=FALSE
  warning = FALSE, # suppress warnings
  message = FALSE, # suppress messages
  error = FALSE    # suppress errors
)
```

Shiny outputs are the dynamic elements of your app that **show results** to the user. Think of them as the *answers* to the *questions* posed by input widgets. The core principle is a simple partnership: every output in the user interface (UI) is paired with a `render` function in the server that creates it.

This connection is what makes Shiny magical. When a user changes an input (like moving a slider), the `render` function automatically re-runs, and the output it's connected to (like a plot or a number) instantly updates.

\newpage

## Example 1: The Multi-Faceted Data Explorer 

This first example demonstrates how a single user input can simultaneously update a **plot**, a **data table**, and a **summary text**. It shows how outputs work together to create a cohesive and informative user experience. We'll use the `palmerpenguins` dataset.

The goal is to let a user select a penguin species and see the corresponding data, a scatter plot of their bill dimensions, and a summary count—all at once.

```{r}
#| title: "Penguin Explorer App"
#| viewerHeight: 800

# --- Load Libraries ---
library(shiny)
library(ggplot2)
library(palmerpenguins)
library(DT)

# --- Define UI ---
ui <- fluidPage(
  titlePanel("Palmer Penguins Explorer"),
  sidebarLayout(
    sidebarPanel(
      # Input: A dropdown to select the penguin species
      selectInput(
        inputId = "species_select",
        label = "Select a Penguin Species:",
        choices = c("Adelie", "Chinstrap", "Gentoo"),
        selected = "Adelie"
      )
    ),
    mainPanel(
      # Output 1: A dynamic summary sentence
      h4(textOutput(outputId = "summary_text")),

      # Output 2: The scatter plot
      plotOutput(outputId = "penguin_plot"),

      # Output 3: An interactive data table
      DTOutput(outputId = "penguin_table")
    )
  )
)

# --- Define Server Logic ---
server <- function(input, output) {

  # Create a reactive expression to filter data based on the input
  # This is efficient because it's calculated once and reused by all outputs
  filtered_data <- reactive({
    penguins[penguins$species == input$species_select, ]
  })

  # Output 1: Render the summary text
  output$summary_text <- renderText({
    # The text string depends on the filtered data
    paste("Showing data for", nrow(filtered_data()), "penguins of the", input$species_select, "species.")
  })

  # Output 2: Render the scatter plot
  output$penguin_plot <- renderPlot({
    ggplot(filtered_data(), aes(x = bill_length_mm, y = bill_depth_mm, color = island)) +
      geom_point(size = 4, alpha = 0.7) +
      labs(
        title = "Bill Length vs. Bill Depth",
        x = "Bill Length (mm)",
        y = "Bill Depth (mm)"
      ) +
      theme_minimal()
  })

  # Output 3: Render the interactive data table
  output$penguin_table <- renderDT({
    datatable(filtered_data())
  })
}

# --- This is the standard way to launch a Shiny app in a .qmd file ---
shinyApp(ui, server)
```

### Why It's Powerful

  * **Reactivity in Action**: Changing the dropdown **instantly** triggers updates across three different types of outputs. The user gets immediate, comprehensive feedback.
  * **Efficiency**: The `reactive()` expression for `filtered_data` is a key concept. The data is filtered only *once* per user change, and that result is then shared efficiently with the plot, table, and text outputs.
  * **Different Views, One Source**: It presents the same underlying data in multiple ways (graphically, textually, and tabularly), giving the user a complete picture.

\newpage

## Example 2: The Interactive World Map 

This example showcases a more advanced and visually impressive output: an interactive map using the `leaflet` package. It demonstrates how a simple dropdown can control a complex, web-based widget.

The goal is to provide a list of cities and have the map instantly pan and zoom to the selected location, placing a marker there.

```{r}
#| title: "Interactive City Map App"
#| viewerHeight: 700

# --- Load Libraries ---
library(shiny)
library(leaflet)

# Create a dataframe of cities and their coordinates for the dropdown
cities_df <- data.frame(
  name = c("San Luis Obispo", "New York", "London", "Tokyo"),
  lat = c(35.2828, 40.7128, 51.5072, 35.6762),
  lng = c(-120.6596, -74.0060, -0.1276, 139.6503),
  zoom = c(11, 10, 10, 9)
)

# --- Define UI ---
ui <- fluidPage(
  titlePanel("Interactive City Map"),
  sidebarLayout(
    sidebarPanel(
      # Input: A dropdown to select a city
      selectInput(
        inputId = "city_select",
        label = "Choose a city to view:",
        choices = cities_df$name
      )
    ),
    mainPanel(
      # Output: The Leaflet map widget
      leafletOutput(outputId = "city_map", height = "600px")
    )
  )
)

# --- Define Server Logic ---
server <- function(input, output) {

  # Output: Render the Leaflet map
  output$city_map <- renderLeaflet({
    # Get the data for the selected city
    selected_city <- cities_df[cities_df$name == input$city_select, ]

    # Create the map. The leaflet() function starts the map,
    # then we chain commands with the pipe (|>).
    leaflet() |>
      addTiles() |> # Adds the base map tiles (like from OpenStreetMap)
      setView( # Sets the map's center and zoom level
        lng = selected_city$lng,
        lat = selected_city$lat,
        zoom = selected_city$zoom
      ) |>
      addMarkers( # Adds a popup marker
        lng = selected_city$lng,
        lat = selected_city$lat,
        popup = paste("Welcome to", selected_city$name, "!")
      )
  })
}

# --- Launch the app ---
shinyApp(ui, server)
```

### Why It's Powerful

  * **Integrating External Libraries**: Shiny's output system is extensible. It seamlessly integrates powerful JavaScript libraries like Leaflet, giving you access to sophisticated web widgets without writing any JavaScript yourself.
  * **High-Impact Visuals**: A dynamic, interactive map is far more engaging than a static image. It allows users to explore spatial data intuitively.
  * **Complex Control, Simple Input**: The user interacts with a simple dropdown, but the result is a complex change in the map's view and markers. This is the essence of a powerful user interface—hiding complexity behind simple controls.
