---
title: "what_is_shiny"
format: docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE,    # don't run code by default
  echo = TRUE,     # still show code unless you hide with echo=FALSE
  warning = FALSE, # suppress warnings
  message = FALSE, # suppress messages
  error = FALSE    # suppress errors
)
```

# Build your first Shiny app in Positron

## 0) Setup (30s)

> “Open Positron. Create or open a folder for the project, e.g., `shiny-starter`. Open a terminal inside Positron. I’m going to check R works and install Shiny.”

```{r}
R.version.string
install.packages("shiny")
```

## 1) Positron specifics (60s)

> “Positron is like VS Code: no ‘Run App’ button (yet). Use the terminal. If preview doesn’t appear, it’ll open in your browser. Use `launch.browser = TRUE` to force it. If packages won’t install, check you’re in the R terminal (not bash) and that your R path is detected. If a port is busy, set `port = 5050` (or another). For larger apps, split into modules under `R/` and static assets under `www/`.”

Recommended layout:

```         
app_name_app/
├─ R/               # (optional) modules, helpers
├─ www/             # (optional) CSS, images
└─ app.R
```

\newpage

## 2) Create Shiny I

**Step 1:** Create a directory named `geyser_app `

**Step 2:** Create an R file name `app.R`,  save the below information in the file

```{r}
library(shiny)

# Define UI for application that draws a histogram
ui <- fluidPage(

    # Application title
    titlePanel("Old Faithful Geyser Data"),

    # Sidebar with a slider input for number of bins 
    sidebarLayout(
        sidebarPanel(
            sliderInput("bins",
                        "Number of bins:",
                        min = 1,
                        max = 50,
                        value = 30)
        ),

        # Show a plot of the generated distribution
        mainPanel(
           plotOutput("distPlot")
        )
    )
)

# Define server logic required to draw a histogram
server <- function(input, output) {

    output$distPlot <- renderPlot({
     geom_histogram(
      binwidth = diff(range(x)) / input$bins,
      fill = "darkgray",
      color = "white"
    )
    })
}

# Run the application 
shinyApp(ui = ui, server = server)

```

**Step 3:** In R Console, run this line `shiny::runApp("geyser_app")`

\newpage

## 3) Create Shiny II

> “Now I’ll create `app.R` with a tiny UI + server. This runs anywhere and shows a reactive histogram.”

Create `app.R` and paste:

```{r}
# app.R — minimal Shiny in Positron
library(shiny)

ui <- fluidPage(
  titlePanel("Hello, Positron + Shiny"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("n", "How many points?", 10, 1000, 200, step = 10)
    ),
    mainPanel(
      plotOutput("hist"),
      verbatimTextOutput("stats")
    )
  )
)

server <- function(input, output, session) {
  x <- reactive(rnorm(input$n))

  output$hist <- renderPlot({
    hist(x(), main = "", xlab = "Value")
  })

  output$stats <- renderPrint({
    summary(x())
  })
}

shinyApp(ui, server)
```

\newpage

> “In Positron, I run the app from the terminal. Positron will show a preview panel or open my browser.”

```{r}
shiny::runApp()                 # add launch.browser = TRUE if you prefer the browser
# shiny::runApp(port = 5050)    # pick a port if needed
```

You should see a histogram that updates as you move the slider.

\newpage

## 4) Make it a bit nicer

> “Let’s add a theme and a data table—useful for teaching demos.”

```{r}
install.packages(c("bslib","DT"))
```

Replace `app.R` with:

```{r}
library(shiny)
library(ggplot2)
library(bslib)
library(DT)

ui <- page_fluid(
  theme = bs_theme(version = 5, bootswatch = "flatly"),
  layout_sidebar(
    sidebar = sidebar(
      sliderInput("bins", "Bins", 5, 60, 20),
      checkboxInput("show_tbl", "Show table", TRUE)
    ),
    card(
      card_header(h4("Reactive Histogram")),
      plotOutput("hist", height = 300),
      verbatimTextOutput("stats"),
      conditionalPanel("input.show_tbl", DTOutput("tbl"))
    )
  )
)

server <- function(input, output, session) {
  x <- reactive(rnorm(300))

output$hist <- renderPlot({
  
  ggplot(data.frame(x = x()), aes(x)) +
    geom_histogram(bins = input$bins, fill = "skyblue", color = "white") +
    labs(x = "Value", y = "Count", title = NULL) +
    theme_minimal(base_size = 14)
})

  output$stats <- renderPrint(summary(x()))

  output$tbl <- renderDT(datatable(data.frame(x = x()), options = list(pageLength = 8)))
}

shinyApp(ui, server)
```

Run again:

```{r}
shiny::runApp()
```

\newpage

## 5) ETV-style tabs for class demos

> “Quick ETV framing: three tabs for Extraction, Transformation, Visualization.”

Replace UI’s main area with:

```{r}
navset_card_pill(
  nav_panel("Extract", verbatimTextOutput("head")),
  nav_panel("Transform", verbatimTextOutput("summary")),
  nav_panel("Visualize", plotOutput("hist2", height = 280))
)
```

…and in server:

```{r}
d <- reactive(iris)

output$head <- renderPrint(head(d()))
output$summary <- renderPrint(summary(d()))
output$hist2 <- renderPlot({
  ggplot(data = d(), aes(x = Sepal.Length)) +
    geom_histogram(binwidth = 0.3, fill = "steelblue", color = "black") +
    labs(x = "Sepal.Length", y = "Frequency")
})
```

