---
title: "ETV Cycle 2 — Who Is the Real GOAT? Reading Multiple Excel Sheets"
format: html
editor: visual
---

## Description

NBA career statistics for Michael Jordan (MJ), LeBron James (LJ), and Kobe Bryant (KB) are stored in a single Excel workbook with one sheet per player. We will extract each sheet, combine the players, and compare their scoring careers.

Each **row** represents one season. Each **column** represents a season-level statistic.

------------------------------------------------------------------------

### Motivating Questions

**Subtask 1** — How does career average points-per-game compare across MJ, LJ, and KB?

**Subtask 2** — How does each player's scoring compare across career stages (Novice, Intermediate, Veteran)?

------------------------------------------------------------------------

## Learning Moment 1 — Libraries

```{r}
library(____)   ## Data Extraction  — file paths
library(____)   ## Data Extraction  — reading Excel (.xlsx) files
library(____)   ## Data Transformation — bind_rows, mutate, group_by, summarize
library(____)   ## Data Transformation — string manipulation
library(____)   ## Data Visualization
library(____)   ## Data Visualization — reordering factor levels in plots
```

**Side Q1.** Why is `readxl` kept separate from `readr` rather than merged into one package?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Extraction (E)

#### Learning Moment 2 — `read_excel()` with `sheet`

The `sheet` argument accepts two forms — **by position** (integer) or **by name** (string):

| Form | Example | Trade-off |
|------------------|-------------------------|-----------------------------|
| Integer | `sheet = 1` | Reads the sheet at that slot; breaks silently if tabs are reordered |
| String | `sheet = "MJ"` | Reads by tab name; self-documenting and survives reordering |

```{r}
## Read each player's sheet from the same workbook
path_to_goat <- here("data", "nba_goat.xlsx") 
mj_df <- read_excel(____, sheet = ____)
lj_df <- read_excel(____, sheet = ____)
kb_df <- read_excel(____, sheet = ____)
```

**Q1.** What does the `sheet` argument in `read_excel()` specify?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

**Side Q2.** How would you list all sheet names in an Excel file without opening it in Excel?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

**Side Q2b.** If a colleague reorders the workbook tabs so "KB" is now the first sheet, which call breaks — `sheet = 3` or `sheet = "KB"`? Why?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Transformation (T)

#### Learning Moment 3 — Adding an Identifier Column and `bind_rows()`

**`mutate()` key idea:** Adds or overwrites columns while keeping all existing columns intact. Each expression is computed row-by-row. Assigning a constant string (e.g., `"Michael Jordan"`) works because R recycles it to fill every row — a concise way to stamp an identifier onto an entire data frame.

```{r}
## Tag each data frame with the player name
mj_df <- mj_df |> mutate(player = ____)
lj_df <- lj_df |> mutate(player = ____)
kb_df <- kb_df |> mutate(player = ____)

## Stack all three into one data frame
nba_df <- ____(mj_df, lj_df, kb_df)

glimpse(nba_df)
```

**Q2.** What does `bind_rows()` do, and why do we add a `player` column first?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

**Side Q3.** What happens if the column names do not match perfectly when using `bind_rows()`?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 4 — Selecting and Renaming Columns

We keep **only the columns needed for Subtask 2**: the player, the season, and points per game.

```{r}
nba_clean_df <- nba_df |>
  select(____, ____, ____) |>
  rename(points_pg = ____)

nba_clean_df
```

**Q3.** What does `rename()` do, and does it modify the data values?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 5 — Career Stages with `stringr` and `case_when()`

Each player's `Season` is stored as text like `"1984-85"`. We extract the starting year with `str_extract()`, order each player's seasons, and label career stages: seasons **1–5 = Novice**, **6–10 = Intermediate**, **11+ = Veteran**.

```{r}
## Label each season with the player's career stage
nba_clean_df <- nba_clean_df |>
  mutate(
    season_num = as.integer(str_extract(Season, "\\d+")),
    experience = case_when(
      season_num <= ____ ~ ____,
      season_nun <= ____ ~ ____,
      TRUE                 ~ ____
    ),
    experience = factor(experience,
                        levels = c(____, ____, ____))
  ) |>
  ungroup()

nba_clean_df
```

**Q4.** What does `case_when()` do, and why does the order of the conditions matter?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

**Side Q4.** Why do we wrap the new `experience` column in `factor()` with explicit `levels`?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 6 — `group_by()` and `summarize()` for Player Comparison

```{r}
avg_ppg_df <- nba_clean_df |>
  group_by(____) |>
  summarize(
    career_avg_ppg = mean(____, na.rm = TRUE),
    sd_ppg         = ____(____, na.rm = TRUE),
    seasons_played = ____(),
    .groups = "drop"
  ) |>
  arrange(desc(____))

avg_ppg_df
```

**Q5.** What does `arrange(desc(career_avg_ppg))` do?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

**Side Q5.** Why do we use `na.rm = TRUE` inside `mean()` and `sd()`?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Visualization (V)

**Style rule:** always pipe the data frame into the plot — `df |> ggplot(aes(...))`. **Never** write `ggplot(df, ...)`.

#### Learning Moment 7 — Bar Chart with Points and Error Bars

```{r}
____ |>
  ggplot(aes(x = fct_reorder(____, ____), y = ____,
             fill = ____)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = ____ - ____,
                    ymax = ____ + ____),
                width = 0.2) +
  coord_flip() +
  labs(
    title = "Career Average PPG — MJ, LJ, KB",
    x     = "Player",
    y     = "Career Avg Points Per Game"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = ____))
```

**Q6.** What does `fct_reorder(player, career_avg_ppg)` do to the bars?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

**Side Q6.** What does `coord_flip()` do, and when is a horizontal bar chart preferable?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 8 — Points and Error Bars Across Career Stages

We now answer Subtask 2 using the career-stage labels from Learning Moment 5: summarize each player's scoring within each stage, then display the **mean as a point** and the **spread as an error bar** (mean ± 1 SD).

```{r}
## Average and spread of PPG per player, per career stage
exp_ppg_df <- nba_clean_df |>
  group_by(____, ____) |>
  summarize(
    avg_ppg = mean(____, na.rm = TRUE),
    sd_ppg  = ____(____, na.rm = TRUE),
    .groups = "drop"
  )

____ |>
  ggplot(aes(x = ____, y = ____, color = ____)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = ____ - ____,
                    ymax = ____ + ____),
                width = 0.2, position = position_dodge(width = 0.5)) +
  labs(
    title = "Points Per Game by Career Stage — MJ, LJ, KB",
    x     = "Career Stage",
    y     = "Average Points Per Game",
    color = "Player"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = ____))
```

**Q7.** What do the error bars represent in this plot, and why are they useful when comparing players?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

### Subtask 3 — Student-Generated Comparison

Propose a question comparing MJ, LJ, and KB on a statistic other than points (e.g., rebounds, assists). Your instructor will select one for class exploration.

Remember the style rule: `df |> ggplot(aes(...))` — never `ggplot(df, ...)`.

```{r}
## Your code here

```