---
title: "ETV Cycle 5 — Wikipedia Automation: `for` Loops & Conditionals"
format: html
editor: visual
---

## Description

In Cycle 3 we scraped one Wikipedia page manually. Here we **automate** that process across multiple state pages using a `for` loop with `if / else` conditionals to handle pages that differ in structure.

------------------------------------------------------------------------

### Motivating Questions

**Subtask 1** — Can we automatically scrape county population data for multiple US states?

**Subtask 2** — How do we handle states whose Wikipedia pages have a different table structure?

------------------------------------------------------------------------

## Learning Moment 1 — Libraries

```{r}
library(____)   ## Data Extraction  — HTML scraping
library(____)   ## Data Transformation
library(____)   ## Data Transformation — string cleaning
library(____)   ## Data Transformation — parse_number()
library(____)   ## Data Transformation — map functions
library(____)   ## Data Visualization
library(____)   ## Data Visualization
```

------------------------------------------------------------------------

## Learning Moment 2 — The `for` Loop

```{r}
states <- c("California", "Texas", "Florida", "New York")

for (____ in ____) {
  cat("Processing:", ____, "\n")
}
```

**Q1.** In the loop above, what is `state` called, and what value does it hold on the third iteration?

<!-- [Place Answer - Begin] -->



<!-- [Place Answer - End] -->

**Side Q1.** What is the difference between `cat()` and `print()` when printing inside a loop?

<!-- [Place Answer - Begin] -->



<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Learning Moment 3 — The `if / else if / else` Conditional

```{r}
classify_region <- function(state) {
  if (state %in% c("California", "Oregon", "Washington")) {
    return(____)
  } else if (state %in% c("New York", "Massachusetts", "Connecticut")) {
    return(____)
  } else if (state %in% c("Texas", "Florida", "Georgia")) {
    return(____)
  } else {
    return(____)
  }
}

classify_region("California")
classify_region("Ohio")
```

**Q2.** What is the role of `else if` in the conditional block above?

<!-- [Place Answer - Begin] -->



<!-- [Place Answer - End] -->

**Side Q2.** What does `return()` do inside a function, and what happens if you omit it?

<!-- [Place Answer - Begin] -->



<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Extraction (E)

#### Learning Moment 4 — Building a URL-Construction Function

```{r}
make_state_url <- function(state_name) {
  base_url    <- "https://en.wikipedia.org/wiki/List_of_counties_in_"
  clean_state <- state_name |>
    ____() |>
    str_replace_all(____, ____)
  str_glue(base_url, clean_state)
}

make_state_url("New Mexico")
```

**Q3.** What does `str_replace_all(" ", "_")` do, and why is it needed for URLs?

<!-- [Place Answer - Begin] -->



<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 5 — The Scraping Loop with a Conditional

```{r}
states_to_scrape <- c("California", "Texas", "Florida",
                       "New York", "Ohio", "Colorado")

all_county_dfs <- ____()   ## empty list to collect results

for (____ in ____) {

  cat("Scraping:", state, "...\n")

  url <- make_state_url(____)

  result <- tryCatch({

    page   <- read_html(____)
    tables <- page |> html_table(fill = TRUE)

    ## Conditional: some pages use a different table index
    if (state %in% c("New York", "Colorado")) {
      tbl_index <- ____
    } else {
      tbl_index <- ____
    }

    df <- tables[[____]] |>
      mutate(state = ____)

    cat("  Success:", nrow(df), "rows\n")
    df

  }, error = function(e) {
    cat("  ERROR for", state, ":", conditionMessage(e), "\n")
    ____   ## return NULL on failure
  })

  all_county_dfs[[____]] <- result

  Sys.sleep(____)   ## polite pause
}
```

**Q4.** What does `tryCatch()` do, and why is it important when scraping multiple pages in a loop?

<!-- [Place Answer - Begin] -->



<!-- [Place Answer - End] -->

**Side Q3.** What does `Sys.sleep(1)` do inside the loop, and why is it considered polite?

<!-- [Place Answer - Begin] -->



<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Transformation (T)

#### Learning Moment 6 — Combining Results with `bind_rows()`

```{r}
## Remove NULL entries and combine
all_county_dfs_clean <- Filter(Negate(is.null), ____)

combined_df <- ____(all_county_dfs_clean)

combined_df |>
  count(____) |>
  arrange(desc(n))
```

**Q5.** What does `Filter(Negate(is.null), all_county_dfs)` do?

<!-- [Place Answer - Begin] -->



<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Visualization (V)

#### Learning Moment 7 — Automating One Plot Per State with a `for` Loop

```{r}
for (____ in unique(combined_df$state)) {

  state_df <- combined_df |>
    filter(state == ____)

  p <- ggplot(state_df, aes(y = fct_reorder(County, Population),
                             x = Population)) +
    geom_col(fill = "steelblue") +
    scale_x_continuous(labels = scales::comma) +
    labs(
      title = paste("County Populations —", ____),
      x     = "Population",
      y     = "County"
    ) +
    theme_minimal(base_size = 8)

  ____(p)
}
```

**Q6.** Why must we call `print(p)` inside a `for` loop to display the plot?

<!-- [Place Answer - Begin] -->



<!-- [Place Answer - End] -->

**Side Q4.** What does `paste()` do in `paste("County Populations —", s)`?

<!-- [Place Answer - Begin] -->



<!-- [Place Answer - End] -->

------------------------------------------------------------------------

### Subtask 3 — Student-Generated Extension

```{r}
## Your code here

```
