---
title: "ETV Cycle 1 — Extracting and Exploring Coffee Purchase Data"
format: html
editor: visual
---

## Description

This dataset contains coffee purchases recorded over several years, beginning in 2018. Each row represents a single transaction capturing the date, amount, and location of the purchase. Coffee spending is a concrete, personal example that makes patterns in time and money immediately meaningful. We will use R to extract this data from a CSV file, transform it into an analysis-ready form, and visualize spending patterns over time.

Each **row** represents a single purchase. Each **column** represents an attribute of that purchase.

------------------------------------------------------------------------

### Motivating Questions

**Subtask 1** — In which months is the most money spent on coffee?

**Subtask 2** — Does this monthly pattern remain consistent across different years?

------------------------------------------------------------------------

## Learning Moment 1 — Packages and Libraries

**Packages** are collections of R functions bundled together to extend R's base capabilities.

**Libraries** are packages that have been loaded into your current R session with `library()`.

###### **Task 1:** Load Libraries

```{r}
library(____)   ## Data Extraction  — file path management
library(____)   ## Data Extraction  — reading CSV files
library(____)   ## Data Transformation — mutate, filter, select, group_by, summarize
library(____)   ## Data Transformation — working with dates
library(____)   ## Data Transformation — reordering factor levels
library(____)   ## Data Visualization  — building plots
```

**Side Q1.** Why do we use `library(dplyr)` instead of `library(tidyverse)` in this course?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Extraction (E)

**Extraction** is the process of bringing data into R from an external source.

**Q1.** Define Data Extraction in your own words.

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 2 — Reading a CSV with `read_csv()`

```{r}
## Read the coffee purchase CSV into R
coffee_df <- ____(here("data", "coffee_purchases.csv"))
```

**Q2.** What does the `<-` operator do in R?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

**Side Q2.** What is the difference between `read_csv()` from `readr` and `read.csv()` from base R?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 3 — Inspecting the Data

###### **Task 2:** Run each inspection function.

```{r}
## Open a spreadsheet-style viewer
____(coffee_df)
```

```{r}
## Check dimensions: rows × columns
____(coffee_df)
```

```{r}
## Print column names and data types
____(coffee_df)
```

**Q3.** What is a data frame (or tibble) in R?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

**Side Q3.** After running `glimpse(coffee_df)`, what does the `<dbl>` type next to the `amount` column tell you?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Transformation (T)

**Transformation** is the process of cleaning, reshaping, and enriching raw data so it is ready for analysis.

**Q4.** Define Data Transformation in your own words.

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 4 — Fixing Dates with `mutate()` and `lubridate`

```{r}
## Convert date column from character to Date class
mod_1_coffee_df <- coffee_df |>
  mutate(date  = ____(date),
         month = month(date, label = ____),
         year  = ____(date))
```

**Q5.** What does `mutate()` do?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

**Side Q4.** Why does `month(date, label = TRUE)` return `"Jan"`, `"Feb"`, etc. instead of `1`, `2`?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 5 — The Pipe Operator `|>`

```{r}
## Select relevant columns using the pipe
coffee_selected_df <- mod_1_coffee_df |>
  select(____, ____, ____, ____, ____)

coffee_selected_df
```

**Q6.** What is the benefit of using `|>` over nesting function calls?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 6 — `filter()`

```{r}
## Keep only purchases of $5 or more
coffee_above5_df <- mod_1_coffee_df |>
  filter(amount ____ 5)
```

```{r}
## Keep only weekend purchases
coffee_weekend_df <- mod_1_coffee_df |>
  filter(day_of_week %in% c(____, ____))
```

**Q7.** What does `%in%` do in the second filter example?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 7 — `group_by()` and `summarize()`

```{r}
## Total and average spending per month
monthly_spending_df <- mod_1_coffee_df |>
  group_by(____) |>
  summarize(
    total_spending = ____(amount),
    avg_spending   = ____(amount),
    n_purchases    = ____(),
    .groups        = "drop"
  )

monthly_spending_df
```

**Q8.** What does `n()` return inside `summarize()`?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

**Side Q5.** What does `.groups = "drop"` do at the end of `summarize()`?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Visualization (V)

**Q9.** Define Data Visualization in your own words.

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 8 — Bar Chart with `ggplot2`

```{r}
ggplot(____, aes(x = ____, y = ____)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Total Coffee Spending by Month",
    x     = "Month",
    y     = "Total Spent ($)"
  ) +
  theme_minimal()
```

**Q10.** What is the role of `aes()` inside `ggplot()`?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

### Subtask 1 — Which months have the highest average coffee spending?

```{r}
ggplot(monthly_spending_df, aes(x = ____, y = ____)) +
  geom_col(fill = "darkorange") +
  labs(
    title = "Average Coffee Spending by Month",
    x     = "Month",
    y     = "Avg Amount Spent ($)"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

### Subtask 2 — Does monthly spending stay consistent across years?

#### Learning Moment 9 — `facet_wrap()`

```{r}
yearly_monthly_df <- mod_1_coffee_df |>
  group_by(____, ____) |>
  summarize(total_spending = sum(amount), .groups = "drop")

p1 <- ggplot(yearly_monthly_df, aes(x = ____, y = ____)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~ ____) +
  labs(
    title = "Monthly Coffee Spending by Year",
    x     = "Month",
    y     = "Total Spent ($)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1
```

**Q11.** What does `facet_wrap(~ year)` do to the plot?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 10 — Customizing with `theme()`

```{r}
p1 +
  theme(
    axis.title.x = element_text(size = ____, face = "bold"),
    axis.title.y = element_text(size = ____, face = "bold"),
    plot.title   = element_text(size = ____, face = "bold", hjust = 0.5)
  )
```

**Q12.** What does `hjust = 0.5` do to the plot title?

<!-- [Place Answer - Begin] -->

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

### Subtask 3 — Student-Generated Question

```{r}
## Your code here

```