---
title: "ETV Cycle 1 — Extracting and Exploring Coffee Purchase Data (SOLUTIONS)"
format: docx
editor: visual
---

## Description

This dataset contains coffee purchases recorded over several years, beginning in 2018. Each row represents a single transaction capturing the date, amount, and location of the purchase. Coffee spending is a concrete, personal example that makes patterns in time and money immediately meaningful. We will use R to extract this data from a CSV file, transform it into an analysis-ready form, and visualize spending patterns over time.

Each **row** represents a single purchase. Each **column** represents an attribute of that purchase.

------------------------------------------------------------------------

### Motivating Questions

**Subtask 1** — In which months is the most money spent on coffee?

**Subtask 2** — Does this monthly pattern remain consistent across different years?

------------------------------------------------------------------------

## Learning Moment 1 — Packages and Libraries

**Packages** are collections of R functions bundled together to extend R's base capabilities.

**Libraries** are packages that have been loaded into your current R session with `library()`.

We load only the packages we need, rather than the entire tidyverse, so you can clearly see which library each function comes from.

###### **Task 1:** Load Libraries

```{r}
library(here)       ## Data Extraction  — file path management
library(readr)      ## Data Extraction  — reading CSV files
library(dplyr)      ## Data Transformation — mutate, filter, select, group_by, summarize
library(lubridate)  ## Data Transformation — working with dates
library(forcats)    ## Data Transformation — reordering factor levels
library(ggplot2)    ## Data Visualization  — building plots
```

**Side Q1.** Why do we use `library(dplyr)` instead of `library(tidyverse)` in this course?

<!-- [Place Answer - Begin] -->

Loading individual libraries (e.g., `library(dplyr)`) forces you to be aware of which package each function belongs to. Loading the full tidyverse loads eight packages at once, masking where functions like `mutate()` or `select()` actually live. Individual loading builds better mental models early in the course.

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Extraction (E)

**Extraction** is the process of bringing data into R from an external source. Raw data lives in files, databases, or online services — R cannot analyze it until it is loaded into your session.

**Q1.** Define Data Extraction in your own words.

<!-- [Place Answer - Begin] -->

Data extraction is the process of importing or retrieving data from an external source — such as a CSV file, database, or API — into R so that it can be inspected and analyzed.

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 2 — Reading a CSV with `read_csv()`

`read_csv()` from the `readr` package reads a comma-separated values file and returns a **tibble** (a modern data frame).

```{r}
## Read the coffee purchase CSV into R
coffee_df <- read_csv(here("ETV", "Data", "coffee_purchases.csv"))
```

Notice the naming convention `coffee_df` — the `_df` suffix signals this object is a data frame.

**Q2.** What does the `<-` operator do in R?

<!-- [Place Answer - Begin] -->

`<-` is the **assignment operator**. It stores the object on the right-hand side into the variable name on the left-hand side. After this line runs, `coffee_df` refers to the loaded data frame throughout the session.

<!-- [Place Answer - End] -->

**Side Q2.** What is the difference between `read_csv()` from `readr` and `read.csv()` from base R?

<!-- [Place Answer - Begin] -->

`read_csv()` (readr) returns a **tibble**, reads faster, never converts strings to factors by default, and provides more informative column-type messages. `read.csv()` (base R) returns a classic `data.frame` and converts character columns to factors by default unless `stringsAsFactors = FALSE` is set.

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 3 — Inspecting the Data

Before transforming any data, always inspect it first to understand its structure, dimensions, and variable types.

###### **Task 2:** Run each of the following inspection functions.

```{r}
## Open a spreadsheet-style viewer
# View(coffee_df)
```

```{r}
## Check dimensions: rows × columns
dim(coffee_df)
```

```{r}
## Print column names and data types
glimpse(coffee_df)
```

**Q3.** What is a data frame (or tibble) in R?

<!-- [Place Answer - Begin] -->

A data frame is a two-dimensional, table-like structure in which each **column** is a variable and each **row** is one observation. Every column can hold a different data type (numeric, character, logical), but all columns must have the same length. A tibble is a modern, enhanced version of a data frame from the tidyverse.

<!-- [Place Answer - End] -->

**Side Q3.** After running `glimpse(coffee_df)`, what does the `<dbl>` type next to the `amount` column tell you?

<!-- [Place Answer - Begin] -->

`<dbl>` stands for **double-precision floating-point number** — R's standard numeric type for decimal values. It tells you the `amount` column stores numbers with decimal places (e.g., 4.75), which is appropriate for dollar amounts.

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Transformation (T)

**Transformation** is the process of cleaning, reshaping, and enriching raw data so it is ready for analysis and visualization.

**Q4.** Define Data Transformation in your own words.

<!-- [Place Answer - Begin] -->

Data transformation is the process of cleaning, reshaping, and modifying a raw dataset — fixing data types, creating new variables, filtering unwanted rows — to produce a tidy, analysis-ready form.

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 4 — Fixing Dates with `mutate()` and `lubridate`

The `date` column likely arrived as a character string rather than a true date. We use `mutate()` from `dplyr` to overwrite it with the correct type using `as_date()` from `lubridate`.

```{r}
## Convert date column from character to Date class
mod_1_coffee_df <- coffee_df |>
  mutate(date  = as_date(date),
         month = month(date, label = TRUE),   ## abbreviated month name (Jan–Dec)
         year  = year(date))
```

**Q5.** What does `mutate()` do?

<!-- [Place Answer - Begin] -->

`mutate()` adds new columns or modifies existing ones **without removing any other columns**. Here it overwrites `date` with a proper Date object and creates two new columns: `month` and `year`, both extracted from `date`.

<!-- [Place Answer - End] -->

**Side Q4.** Why does `month(date, label = TRUE)` return `"Jan"`, `"Feb"`, etc. instead of `1`, `2`?

<!-- [Place Answer - Begin] -->

Setting `label = TRUE` tells lubridate to return an **ordered factor** of abbreviated month names instead of the integer (1–12). Ordered factors preserve calendar order in plots and summaries automatically, so January always appears before February even if the data is unsorted.

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 5 — The Pipe Operator `|>`

The native pipe `|>` passes the left-hand result directly into the first argument of the right-hand function, allowing operations to be chained in readable, top-to-bottom order.

```{r}
## Using the pipe to chain: select relevant columns
coffee_selected_df <- mod_1_coffee_df |>
  select(date, month, year, amount, description)

coffee_selected_df
```

**Q6.** What is the benefit of using `|>` over nesting function calls?

<!-- [Place Answer - Begin] -->

The pipe produces code that reads like a recipe — each step is on its own line, in the order it executes. Nested calls like `select(mutate(coffee_df, ...), ...)` must be read from the inside out, which is harder to follow and debug.

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 6 — `filter()`

`filter()` keeps only the rows where a specified condition is `TRUE`.

```{r}
## Keep only purchases of $5 or more
coffee_above5_df <- mod_1_coffee_df |>
  filter(amount >= 5)
```

```{r}
## Keep only weekend purchases
coffee_weekend_df <- mod_1_coffee_df |>
  filter(day_of_week %in% c("Saturday", "Sunday"))
```

**Q7.** What does `%in%` do in the second filter example?

<!-- [Place Answer - Begin] -->

`%in%` is a membership operator. `day_of_week %in% c("Saturday", "Sunday")` returns `TRUE` for any row where `day_of_week` matches either `"Saturday"` or `"Sunday"`, and `FALSE` otherwise. It is a concise alternative to writing `day_of_week == "Saturday" | day_of_week == "Sunday"`.

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 7 — `group_by()` and `summarize()`

`group_by()` splits the data into groups; `summarize()` collapses each group into one summary row.

```{r}
## Total and average spending per month
monthly_spending_df <- mod_1_coffee_df |>
  group_by(month) |>
  summarize(
    total_spending = sum(amount),
    avg_spending   = mean(amount),
    n_purchases    = n(),
    .groups        = "drop"
  )

monthly_spending_df
```

**Q8.** What does `n()` return inside `summarize()`?

<!-- [Place Answer - Begin] -->

`n()` counts the **number of rows** in each group. Here it tells us how many coffee purchases were made in each month, giving context to the total and average spending figures.

<!-- [Place Answer - End] -->

**Side Q5.** What does `.groups = "drop"` do at the end of `summarize()`?

<!-- [Place Answer - Begin] -->

After `summarize()`, dplyr retains the grouping structure by default (minus the last grouping variable). `.groups = "drop"` explicitly removes all grouping from the resulting data frame, returning a plain tibble. This prevents unexpected behavior in any subsequent operations that happen to depend on group structure.

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

## Visualization (V)

**Visualization** is the process of creating graphical representations of data to reveal patterns and communicate insights.

**Q9.** Define Data Visualization in your own words.

<!-- [Place Answer - Begin] -->

Data visualization is the use of charts, graphs, and other graphical displays to communicate patterns and relationships in data in a way that is faster and more intuitive than reading raw numbers.

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 8 — Bar Chart with `ggplot2`

`ggplot2` builds visualizations in layers using the `+` operator. Each layer adds a visual element to the plot.

```{r}
ggplot(monthly_spending_df, aes(x = month, y = total_spending)) +
  geom_col(fill = "steelblue") +
  labs(
    title   = "Total Coffee Spending by Month",
    x       = "Month",
    y       = "Total Spent ($)"
  ) +
  theme_minimal()
```

**Q10.** What is the role of `aes()` inside `ggplot()`?

<!-- [Place Answer - Begin] -->

`aes()` (aesthetics) maps variables in the data frame to visual properties of the plot. Here `x = month` places month on the horizontal axis and `y = total_spending` controls bar height. Everything inside `aes()` is data-driven; fixed visual properties (like `fill = "steelblue"`) go outside `aes()`.

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

### Subtask 1 — Which months have the highest average coffee spending?

```{r}
ggplot(monthly_spending_df, aes(x = month, y = avg_spending)) +
  geom_col(fill = "darkorange") +
  labs(
    title = "Average Coffee Spending by Month",
    x     = "Month",
    y     = "Avg Amount Spent ($)"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

### Subtask 2 — Does monthly spending stay consistent across years?

#### Learning Moment 9 — `facet_wrap()`

`facet_wrap()` creates a separate subplot for each level of a grouping variable — here, each year gets its own panel.

```{r}
## Year-over-year monthly spending using facets
yearly_monthly_df <- mod_1_coffee_df |>
  group_by(year, month) |>
  summarize(total_spending = sum(amount), .groups = "drop")

p1 <- ggplot(yearly_monthly_df, aes(x = month, y = total_spending)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~ year) +
  labs(
    title = "Monthly Coffee Spending by Year",
    x     = "Month",
    y     = "Total Spent ($)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1
```

**Q11.** What does `facet_wrap(~ year)` do to the plot?

<!-- [Place Answer - Begin] -->

`facet_wrap(~ year)` splits the visualization into one panel per unique value of `year`. Each panel shows the same type of chart (monthly spending bars) but for a single year, making it easy to compare whether the monthly pattern is stable or shifts across years.

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

#### Learning Moment 10 — Customizing with `theme()`

```{r}
p1 +
  theme(
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 12, face = "bold"),
    plot.title   = element_text(size = 15, face = "bold", hjust = 0.5)
  )
```

**Q12.** What does `hjust = 0.5` do to the plot title?

<!-- [Place Answer - Begin] -->

`hjust` controls **horizontal justification**. A value of `0.5` centers the title over the plot area. `hjust = 0` aligns the title to the left; `hjust = 1` aligns it to the right.

<!-- [Place Answer - End] -->

------------------------------------------------------------------------

### Subtask 3 — Student-Generated Question

Propose a question involving both the `description` variable and a time component (month, year, or day of week). Your instructor will select one to explore as a class.

```{r}
## Your code here

```