---
title: "Videos 9 - 12 (EMPTY)"
format: docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## **Video** 9) Iterations – 1: Looping Structures

Looping structures are used to repeat a block of code multiple times.

### **`for` loop**:

Repeats code for each item in a sequence. This is useful when you have a known number of iterations.

```{r}
# Get the mean of a few columns in mtcars
cols_to_summarize <- c("mpg", "hp", "wt")

for (col in cols_to_summarize) {
  # Calculate the mean of the current column
  current_mean <- mean(mtcars[[col]])
  # Print the result
  print(paste("Mean of", col, ":", round(current_mean, 2)))
}
```

#### Question 1

Write a `for` loop that iterates through the sequence of numbers from 1 to 5, printing a descriptive message for each iteration.

```{r}
##----------
## FILLED
##----------
# This loop will run for each number in the sequence from 1 to 5
for (i in 1:5) {
  # The 'i' variable will hold the current number in the sequence (1, then 2, etc.)
  print(paste("This is iteration number:", i))
}
```

\newpage


### **`while` loop**:

Repeats code as long as a specified condition is **TRUE**. This is useful when the number of iterations is not known beforehand.

```{r}
# Print car names until horsepower (hp) exceeds 120
i <- 1
# The loop continues as long as the hp of the car at row 'i' is less than 120
while (mtcars$hp[i] < 120) {
  print(rownames(mtcars)[i])
  i <- i + 1
}
```

#### Question 2

Using a `while` loop, write a script that counts from 1 to 5, printing a message like 'The current count is: \[number\]' at each step.

```{r}
##----------
## FILLED
##----------
# 1. Initialize a counter variable before the loop starts
counter <- 1

# 2. Set the condition. The loop will run as long as 'counter' is less than or equal to 5.
while (counter <= 5) {
  
  # 3. This is the code that gets executed in each iteration
  print(paste("The current count is:", counter))
  
  # 4. IMPORTANT: Increment the counter. Without this, the loop would run forever.
  counter <- counter + 1
}
```

\newpage


### **`repeat` loop**:

Repeats code indefinitely until a `break` statement is executed.

```{r}
# Find the first car with 8 cylinders
i <- 1
repeat {
  # Check if the current car has 8 cylinders
  if (mtcars$cyl[i] == 8) {
    print(paste("First 8-cylinder car found:", rownames(mtcars)[i]))
    break # Exit the loop
  }
  i <- i + 1
}
```

#### Question 3

Write a `repeat` loop that iterates through the `mtcars` data frame row by row to find the first car with exactly 8 cylinders (`cyl`). When it finds the car, it should print a message like "The first 8-cylinder car is: \[Car Name\]" and then stop immediately.

```{r}
##----------
## FILLED
##----------
# 1. Initialize a counter to start at the first row
i <- 1

# 2. Start the repeat loop, which will run indefinitely without a 'break'
repeat {
  
  # 3. Check if the condition is met for the current row
  if (mtcars$cyl[i] == 8) {
    
    # 4. If the condition is TRUE, print the message
    print(paste("The first 8-cylinder car is:", rownames(mtcars)[i]))
    
    # 5. IMPORTANT: Use 'break' to exit the loop immediately
    break
  }
  
  # 6. If the condition is FALSE, increment the counter to check the next row
  i <- i + 1
}
```

## \[END\] Video 9

\newpage


\newpage


## **Video** 10) Control Structures: Conditional Logic

Control structures allow you to execute different blocks of code based on specified conditions.

### **`if`, `else if`, `else`**:

Executes code based on a series of logical checks.

```{r}
# Check the fuel efficiency of the 'Mazda RX4'
mpg <- mtcars["Mazda RX4", "mpg"]

if (mpg > 25) {
  print("High fuel efficiency")
} else if (mpg >= 18) {
  print("Average fuel efficiency")
} else {
  print("Low fuel efficiency")
}
```

#### Question 1

Write a script that first creates a variable named `score` and assigns it the value `85`. Then, using an `if`, `else if`, and `else` structure, print the corresponding letter grade: "Grade: A" for scores 90 and above, "Grade: B" for scores 80 and above, and "Grade: C or below" for all other scores.

```{r}
##----------
## FILLED
##----------
# Create a variable for a student's score
score <- 85

if (score >= 90) {
  print("Grade: A")
} else if (score >= 80) {
  print("Grade: B")
} else {
  print("Grade: C or below")
}

```

\newpage


### **`switch()`**:

Selects a block of code to execute based on an exact match of a value. It's a cleaner alternative to a long `if/else if` chain.

```{r}
# Describe a car based on its number of cylinders
cylinders <- mtcars["Valiant", "cyl"]

switch(as.character(cylinders),
       "4" = "A small, efficient engine.",
       "6" = "A balanced, mid-size engine.",
       "8" = "A large, powerful engine.",
       "Unknown engine size."
)
```

#### Question 2

Write a script that first creates a variable named `day` and assigns it the value "Wednesday". Then, use the `switch()` function to print a specific message if the day is "Monday" or "Friday", and a default message for any other day of the week.

```{r}
##----------
## FILLED
##----------
# Create a variable for a day of the week
day <- "Wednesday"

switch(day,
       "Monday" = "It's the start of the week.",
       "Friday" = "It's the end of the week!",
       "It's a day in the middle of the week." # Default case
)

```

## \[END\] Video 10

\newpage


\newpage


## **Video** 11) Functions: Defining and Using

Functions are reusable blocks of code that perform a specific task.

### **Basic function structure**:

Defines a simple function that takes inputs (arguments) and provides an output (`return` value).

```{r}
# Create a function to convert weight from (lbs/1000) to kilograms
convert_wt_to_kg <- function(weight_in_tons) {
  kg <- weight_in_tons * 1000 * 0.453592
  return(kg)
}

# Use the function on the 'Datsun 710'
datsun_kg <- convert_wt_to_kg(mtcars["Datsun 710", "wt"])
print(datsun_kg)
```

#### Question 1

Write a custom R function named `add_numbers` that accepts two arguments, `x` and `y`. Inside the function, calculate their sum and use the `return()` statement to output the result. After defining the function, test it by calling it with the numbers `5` and `10`, and then print the final result to the console.

```{r}
##----------
## FILLED
##----------
# Create a function that adds two numbers
add_numbers <- function(x, y) {
  result <- x + y
  return(result)
}

# Use the function
sum_result <- add_numbers(5, 10)
print(sum_result)

```

\newpage


### **Function with default arguments**:

Assigns a default value to an argument, which is used if no value is provided by the user.

```{r}
# Function to get the top 'n' cars based on a column
get_top_cars <- function(n = 3, column = "hp") {
  # Order the data frame by the specified column in descending order
  sorted_cars <- mtcars[order(mtcars[[column]], decreasing = TRUE), ]
  # Return the top 'n' rows
  return(head(sorted_cars, n))
}

# Call with default arguments (top 3 by horsepower)
get_top_cars()
```

#### Question 2

Write a custom R function named `greet` that takes a single argument, `name`. Set a default value for `name` to be "Guest". The function should combine the string "Hello, " with the provided name and an exclamation mark. After defining the function, test it twice: first by calling it with no arguments to see the default behavior, and a second time by providing the name "Ana".

```{r}
##----------
## FILLED
##----------

# A function to create a greeting message
greet <- function(name = "Guest") {
  paste("Hello,", name, "!")
}

# Call with the default argument
greet()
# Call with a provided argument
greet("Ana")

```

\newpage


### **Multiple outputs (as a list)**:

A function can return multiple values by packaging them into a list.

```{r}
# Function to get key stats for a column
get_stats <- function(column_vector) {
  list(
    average = mean(column_vector),
    median_val = median(column_vector),
    std_dev = sd(column_vector)
  )
}

# Get stats for the mpg column
mpg_stats <- get_stats(mtcars$mpg)
print(mpg_stats$average)
```

#### Question 3

Write a custom R function named `get_vector_stats` that accepts a single argument, a numeric vector. Inside the function, calculate both the `mean` and the `sum` of the vector and store them as a named `list`. The function should then `return` this list. After defining the function, test it by calling it with the vector `c(10, 20, 30)` and printing only the `average` from the resulting list.

```{r}
##----------
## FILLED
##----------

# Function to get key stats for a numeric vector
get_vector_stats <- function(num_vector) {
  stat_list <- list(
    average = mean(num_vector),
    total_sum = sum(num_vector)
  )
  return(stat_list)
}

# Get stats for a simple vector
stats <- get_vector_stats(c(10, 20, 30))
print(stats$average)

```

## \[END\] Video 11

\newpage


\newpage


## **Video** 12) Iterations – 2: Functional Programming with `{purrr}`

In this video, we'll focus on functional programming, a powerful alternative to writing loops. We will use the **`{purrr}`** package, which provides a consistent set of tools for iterating over lists and vectors. The main idea is to apply a function to each element of an object. This often results in code that is more concise and easier to read than a `for` loop.

***#Fill in the blank***

```{r}
##----------
## FILLED
##----------
# Load the purrr package
library(purrr)
```

\newpage


### **`map()`: Apply a function to each element**

The `map()` function is the foundation of `{purrr}`. It takes a vector or list as input, applies a function to each of its elements, and **always returns a list**.

***#Fill in the blank***

```{r}
##----------
## FILLED
##----------
# Let's create a simple function to square a number
square_a_number <- function(x) {
  return(x^2)
}

# Now, use map() to apply this function to each number in our vector
numbers_to_square <- c(1, 2, 3, 4)

# map() will return a list of the results
squared_list <- map(numbers_to_square, square_a_number)

print(squared_list)
```

### **Type-Specific Maps: `map_dbl()` and `map_chr()`**

Often, you don't want a list back; you want a simple vector. `purrr` provides variants like `map_dbl()` (returns a numeric/double vector) and `map_chr()` (returns a character vector) for this purpose.

***#Fill in the blank***

```{r}
##----------
## FILLED
##----------

# Using map_dbl() returns a clean numeric vector instead of a list
squared_vector <- map_dbl(numbers_to_square, square_a_number)

print(squared_vector)
```

\newpage


```{r}
##----------
## FILLED
##----------
# Let's create a function that returns text
create_greeting <- function(name) {
  paste("Hello,", name)
}

student_names <- c("Ana", "Ben", "Charlie")

# Use map_chr() to get a character vector of greetings
greetings <- map_chr(student_names, create_greeting)

print(greetings)
```

\newpage


### **`map2()`: Iterate over two vectors at once**

The `map2()` function lets you iterate over two vectors in parallel. The function you provide must accept two arguments.

***#Fill in the new blank***

```{r}
##----------
## FILLED
##----------


# Let's create a simple function to add two numbers
add_two_numbers <- function(x, y) {
  return(x + y)
}

# Create two vectors of the same length
first_numbers <- c(10, 20, 30)
second_numbers <- c(1, 2, 3)

# Use map2_dbl() to apply the function to corresponding elements
# The first element of 'first_numbers' is added to the first of 'second_numbers', and so on.
sums <- map2_dbl(first_numbers, second_numbers, add_two_numbers)

print(sums)
```

\newpage


### **`walk()`: For functions with side effects**

Sometimes you want to call a function not for what it *returns*, but for what it *does*—like printing a message or saving a file. This is called a "side effect." `walk()` works just like `map()`, but it doesn't return anything. It's used exclusively for triggering side effects.

***#Fill in the blank***

#### Method 1

```{r}
##----------
## FILLED
##----------


# Let's create a function that just prints a message
print_a_summary <- function(student, grade) {
  print(paste(student, "received a grade of", grade))
}

# Create our data
student_data <- data.frame(
  name = c("Ana", "Ben", "Charlie"),
  final_grade = c(92, 85, 88)
)

# Use walk2() to iterate over two columns and print a summary for each student
# Note that nothing is saved to an object. The output is printed directly to the console.
walk2(student_data$name, student_data$final_grade, print_a_summary)
```

#### Method 2

```{r}
##----------
## FILLED
##----------


# Function and data from before
print_a_summary <- function(student, grade) {
  print(paste(student, "received a grade of", grade))
}

student_data <- data.frame(
  name = c("Ana", "Ben", "Charlie"),
  final_grade = c(92, 85, 88)
)

# Pipe the data frame into pwalk()
student_data %>%
  pwalk(~print_a_summary(student = .x, grade = .y))
```

## \[END\] Video 12

\newpage


\newpage

