---
title: "video_12"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


## **Video** 12) Iterations – 2: Functional Programming with `{purrr}`

In this video, we'll focus on functional programming, a powerful alternative to writing loops. We will use the **`{purrr}`** package, which provides a consistent set of tools for iterating over lists and vectors. The main idea is to apply a function to each element of an object. This often results in code that is more concise and easier to read than a `for` loop.

***#Fill in the blank***

```{r}
##----------
## EMPTY
##----------
# Load the purrr package
____(____)
```

::: {style="height: 300px; background-color: white;"}
:::

### **`map()`: Apply a function to each element**

The `map()` function is the foundation of `{purrr}`. It takes a vector or list as input, applies a function to each of its elements, and **always returns a list**.

***#Fill in the blank***

```{r}
##----------
## EMPTY
##----------
# Let's create a simple function to square a number
square_a_number <- function(____) {
return(____^____)
}

# Now, use map() to apply this function to each number in our vector
numbers_to_square <- c(____, ____, ____, ____)

# map() will return a list of the results
squared_list <- map(____, ____)

print(____)
```

### **Type-Specific Maps: `map_dbl()` and `map_chr()`**

Often, you don't want a list back; you want a simple vector. `purrr` provides variants like `map_dbl()` (returns a numeric/double vector) and `map_chr()` (returns a character vector) for this purpose.

***#Fill in the blank***

```{r}
##----------
## EMPTY
##----------

# Using map_dbl() returns a clean numeric vector instead of a list
squared_vector <- map_dbl(____, ____)

print(____)
```

::: {style="height: 300px; background-color: white;"}
:::

```{r}
##----------
## EMPTY
##----------
# Let's create a function that returns text
create_greeting <- function(____) {
paste("Hello,", ____)
}

student_names <- c(____, ____, ____)

# Use map_chr() to get a character vector of greetings
greetings <- map_chr(____, ____)

print(____)
```

::: {style="height: 300px; background-color: white;"}
:::

### **`map2()`: Iterate over two vectors at once**

The `map2()` function lets you iterate over two vectors in parallel. The function you provide must accept two arguments.

***#Fill in the new blank***

```{r}
##----------
## EMPTY
##----------


# Let's create a simple function to add two numbers
add_two_numbers <- function(____, ____) {
return(____ + ____)
}

# Create two vectors of the same length
first_numbers <- c(____, ____, ____)
second_numbers <- c(____, ____, ____)

# Use map2_dbl() to apply the function to corresponding elements
# The first element of 'first_numbers' is added to the first of 'second_numbers', and so on.
sums <- map2_dbl(____, ____, ____)

print(____)
```

::: {style="height: 300px; background-color: white;"}
:::

### **`walk()`: For functions with side effects**

Sometimes you want to call a function not for what it *returns*, but for what it *does*—like printing a message or saving a file. This is called a "side effect." `walk()` works just like `map()`, but it doesn't return anything. It's used exclusively for triggering side effects.

***#Fill in the blank***

#### Method 1

```{r}
##----------
## EMPTY
##----------


# Let's create a function that just prints a message
print_a_summary <- function(____, ____) {
print(paste(____, "received a grade of", ____))
}

# Create our data
student_data <- data.frame(
name = c("Ana", "Ben", "Charlie"),
final_grade = c(92, 85, 88)
)

# Use walk2() to iterate over two columns and print a summary for each student
# Note that nothing is saved to an object. The output is printed directly to the console.
walk2(____, ____, ____)
```

#### Method 2

```{r}
##----------
## EMPTY
##----------


# Function and data from before
print_a_summary <- function(____, ____) {
print(paste(____, "received a grade of", ____))
}

student_data <- data.frame(
name = c("Ana", "Ben", "Charlie"),
final_grade = c(92, 85, 88)
)

# Pipe the data frame into pwalk()
student_data %>%
pwalk(~print_a_summary(student = ____, grade = ____))
```

## \[END\] Video 12

::: {style="height: 300px; background-color: white;"}
:::

::: {style="height: 300px; background-color: white;"}
:::
