---
title: "video_10"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


## **Video** 10) Control Structures: Conditional Logic

Control structures allow you to execute different blocks of code based on specified conditions.

### **`if`, `else if`, `else`**:

Executes code based on a series of logical checks.

```{r}
# Check the fuel efficiency of the 'Mazda RX4'
mpg <- mtcars["Mazda RX4", "mpg"]

if (mpg > 25) {
print("High fuel efficiency")
} else if (mpg >= 18) {
print("Average fuel efficiency")
} else {
print("Low fuel efficiency")
}
```

#### Question 1

Write a script that first creates a variable named `score` and assigns it the value `85`. Then, using an `if`, `else if`, and `else` structure, print the corresponding letter grade: "Grade: A" for scores 90 and above, "Grade: B" for scores 80 and above, and "Grade: C or below" for all other scores.

```{r}
##----------
## EMPTY
##----------
# Create a variable for a student's score
score <- ____

if (____ >= ____) {
print("Grade: A")
} else if (____ >= ____) {
print("Grade: B")
} else {
print("Grade: C or below")
}

```

::: {style="height: 300px; background-color: white;"}
:::

### **`switch()`**:

Selects a block of code to execute based on an exact match of a value. It's a cleaner alternative to a long `if/else if` chain.

```{r}
# Describe a car based on its number of cylinders
cylinders <- mtcars["Valiant", "cyl"]

switch(as.character(cylinders),
"4" = "A small, efficient engine.",
"6" = "A balanced, mid-size engine.",
"8" = "A large, powerful engine.",
"Unknown engine size."
)
```

#### Question 2

Write a script that first creates a variable named `day` and assigns it the value "Wednesday". Then, use the `switch()` function to print a specific message if the day is "Monday" or "Friday", and a default message for any other day of the week.

```{r}
##----------
## EMPTY
##----------
# Create a variable for a day of the week
day <- ____

switch(____,
"Monday" = ____,
"Friday" = ____,
____ # Default case
)

```

## \[END\] Video 10

::: {style="height: 300px; background-color: white;"}
:::

::: {style="height: 300px; background-color: white;"}
:::

