---
title: "video_11"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


## **Video** 11) Functions: Defining and Using

Functions are reusable blocks of code that perform a specific task.

### **Basic function structure**:

Defines a simple function that takes inputs (arguments) and provides an output (`return` value).

```{r}
# Create a function to convert weight from (lbs/1000) to kilograms
convert_wt_to_kg <- function(weight_in_tons) {
kg <- weight_in_tons * 1000 * 0.453592
return(kg)
}

# Use the function on the 'Datsun 710'
datsun_kg <- convert_wt_to_kg(mtcars["Datsun 710", "wt"])
print(datsun_kg)
```

#### Question 1

Write a custom R function named `add_numbers` that accepts two arguments, `x` and `y`. Inside the function, calculate their sum and use the `return()` statement to output the result. After defining the function, test it by calling it with the numbers `5` and `10`, and then print the final result to the console.

```{r}
##----------
## EMPTY
##----------
# Create a function that adds two numbers
add_numbers <- function(____, ____) {
result <- ____ + ____
return(____)
}

# Use the function
sum_result <- add_numbers(____, ____)
print(____)

```

::: {style="height: 300px; background-color: white;"}
:::

### **Function with default arguments**:

Assigns a default value to an argument, which is used if no value is provided by the user.

```{r}
# Function to get the top 'n' cars based on a column
get_top_cars <- function(n = 3, column = "hp") {
# Order the data frame by the specified column in descending order
sorted_cars <- mtcars[order(mtcars[[column]], decreasing = TRUE), ]
# Return the top 'n' rows
return(head(sorted_cars, n))
}

# Call with default arguments (top 3 by horsepower)
get_top_cars()
```

#### Question 2

Write a custom R function named `greet` that takes a single argument, `name`. Set a default value for `name` to be "Guest". The function should combine the string "Hello, " with the provided name and an exclamation mark. After defining the function, test it twice: first by calling it with no arguments to see the default behavior, and a second time by providing the name "Ana".

```{r}
##----------
## EMPTY
##----------

# A function to create a greeting message
greet <- function(name = ____) {
paste("Hello,", ____, "!")
}

# Call with the default argument
greet(____)
# Call with a provided argument
greet(____)

```

::: {style="height: 300px; background-color: white;"}
:::

### **Multiple outputs (as a list)**:

A function can return multiple values by packaging them into a list.

```{r}
# Function to get key stats for a column
get_stats <- function(column_vector) {
list(
average = mean(column_vector),
median_val = median(column_vector),
std_dev = sd(column_vector)
)
}

# Get stats for the mpg column
mpg_stats <- get_stats(mtcars$mpg)
print(mpg_stats$average)
```

#### Question 3

Write a custom R function named `get_vector_stats` that accepts a single argument, a numeric vector. Inside the function, calculate both the `mean` and the `sum` of the vector and store them as a named `list`. The function should then `return` this list. After defining the function, test it by calling it with the vector `c(10, 20, 30)` and printing only the `average` from the resulting list.

```{r}
##----------
## EMPTY
##----------

# Function to get key stats for a numeric vector
get_vector_stats <- function(____) {
stat_list <- list(
average = ____(____),
total_sum = ____(____)
)
return(____)
}

# Get stats for a simple vector
stats <- get_vector_stats(c(____, ____, ____))
print(____$____)

```

## \[END\] Video 11

::: {style="height: 300px; background-color: white;"}
:::

::: {style="height: 300px; background-color: white;"}
:::

