---
title: "video_9"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## **Video** 9) Iterations – 1: Looping Structures

Looping structures are used to repeat a block of code multiple times.

### **`for` loop**:

Repeats code for each item in a sequence. This is useful when you have a known number of iterations.

```{r}
# Get the mean of a few columns in mtcars
cols_to_summarize <- c("mpg", "hp", "wt")

for (col in cols_to_summarize) {
# Calculate the mean of the current column
current_mean <- mean(mtcars[[col]])
# Print the result
print(paste("Mean of", col, ":", round(current_mean, 2)))
}
```

#### Question 1

Write a `for` loop that iterates through the sequence of numbers from 1 to 5, printing a descriptive message for each iteration.

```{r}
##----------
## EMPTY
##----------
# This loop will run for each number in the sequence from 1 to 5
for (____ in ____) {
# The 'i' variable will hold the current number in the sequence (1, then 2, etc.)
print(paste("This is iteration number:", ____))
}
```

::: {style="height: 300px; background-color: white;"}
:::

### **`while` loop**:

Repeats code as long as a specified condition is **TRUE**. This is useful when the number of iterations is not known beforehand.

```{r}
# Print car names until horsepower (hp) exceeds 120
i <- 1
# The loop continues as long as the hp of the car at row 'i' is less than 120
while (mtcars$hp[i] < 120) {
print(rownames(mtcars)[i])
i <- i + 1
}
```

#### Question 2

Using a `while` loop, write a script that counts from 1 to 5, printing a message like 'The current count is: \[number\]' at each step.

```{r}
##----------
## EMPTY
##----------
# 1. Initialize a counter variable before the loop starts
counter <- ____

# 2. Set the condition. The loop will run as long as 'counter' is less than or equal to 5.
while (____ <= ____) {

# 3. This is the code that gets executed in each iteration
print(paste("The current count is:", ____))

# 4. IMPORTANT: Increment the counter. Without this, the loop would run forever.
counter <- ____ + ____
}
```

::: {style="height: 300px; background-color: white;"}
:::

### **`repeat` loop**:

Repeats code indefinitely until a `break` statement is executed.

```{r}
# Find the first car with 8 cylinders
i <- 1
repeat {
# Check if the current car has 8 cylinders
if (mtcars$cyl[i] == 8) {
print(paste("First 8-cylinder car found:", rownames(mtcars)[i]))
break # Exit the loop
}
i <- i + 1
}
```

## \[END\] Video 9

::: {style="height: 300px; background-color: white;"}
:::

::: {style="height: 300px; background-color: white;"}
:::





