---
title: "Videos 5 - 8 (FILLED)"
format: docx
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

### **Video 5) `stringr`: Working with Text Data**

#### **Video Notes**

In this video, we'll explore the `stringr` package, which provides a consistent and easy-to-use set of tools for working with character strings. We'll load the `stringr` package and create a sample vector of text to work with for our examples.

***#Fill in the blank***

```{r}
# Load the stringr package
library(stringr)

# Create a vector of sample sentences
sentences <- c(
  "The quick brown fox jumps over the lazy dog.",
  "R is a language for statistical computing.",
  "The year is 2025, and the sky is blue.",
  "Another sentence with the word dog."
)
```

\newpage

#### **`str_detect()`: Detect a Pattern**

`str_detect()` returns a logical (TRUE/FALSE) vector indicating whether a pattern is found in each string.

***#Fill in the blank***

```{r}
# Detect which sentences contain the word "dog"
str_detect(sentences, pattern = "dog")
```

\newpage

#### **`str_replace_all()`: Replace Patterns**

`str_replace_all()` finds all matches of a pattern and replaces them with a new string.

***#Fill in the blank***

```{r}
# Replace all instances of the word "dog" with "cat"
str_replace_all(sentences, pattern = "dog", replacement = "cat")
```

\newpage

#### **`str_to_upper()`: Change Case**

Functions like `str_to_upper()`, `str_to_lower()`, and `str_to_title()` change the case of your strings.

***#Fill in the blank***

```{r}
# Convert the first sentence to all uppercase
str_to_upper(sentences[1])
```

\newpage

#### **`str_subset()`: Keep Strings with a Pattern**

`str_subset()` is a useful shortcut that returns only the strings from your vector that contain a match for the pattern.

***#Fill in the blank***

```{r}
# Keep only the sentences that contain the word "dog"
str_subset(sentences, pattern = "dog")
```

\newpage

#### **`str_extract()`: Extract a Pattern**

`str_extract()` pulls out the first occurrence of a pattern. This is powerful when combined with regular expressions to find specific types of text, like numbers. The pattern `\\d+` matches one or more digits.

***#Fill in the blank***

```{r}
# Extract the number from our sentences vector
str_extract(sentences, pattern = "\\d+")
```

\newpage

#### **`str_split()`: Split a String**

`str_split()` breaks a string into a list of smaller pieces based on a delimiter pattern.

***#Fill in the blank***

```{r}
# Split the first sentence into a vector of individual words
# The pattern " " splits the string at each space
str_split(sentences[1], pattern = " ")
```

------------------------------------------------------------------------

## \[END\] Video 5

\newpage

\newpage

### **Video 6) `forcats`: Working with Factors**

#### **Video Notes**

This video covers the `forcats` package, which is designed for working with factors. Factors are R's way of handling categorical variables, and `forcats` provides tools to solve common problems, like changing the order of levels or re-grouping them.

***#Fill in the blank***

```{r}
# Load the forcats package
library(forcats)

# Create a sample factor representing survey responses
responses <- factor(c("Agree", "Disagree", "Agree", "Neutral", "Disagree", "Agree"))
```

\newpage

#### **`fct_relevel()`: Manually Reorder Levels**

By default, factor levels are ordered alphabetically. `fct_relevel()` lets you move specific levels to the front, changing their order for plots and tables.

***#Fill in the blank***

```{r}
# The original order is alphabetical: "Agree", "Disagree", "Neutral"
levels(responses)

# Move "Neutral" to be the second level, after "Agree"
fct_relevel(responses, "Neutral", after = 1)
```

\newpage

#### **`fct_infreq()`: Reorder by Frequency**

`fct_infreq()` is a useful shortcut to reorder factor levels so that the most common level appears first.

***#Fill in the blank***

```{r}
# Reorder the levels based on how many times each one appears
fct_infreq(responses)
```

\newpage

#### **`fct_rev()`: Reverse Level Order**

`fct_rev()` simply reverses the current order of the levels.

***#Fill in the blank***

```{r}
# Reverse the order of the levels in our responses factor
fct_rev(responses)
```

\newpage

#### **`fct_recode()`: Change Level Values**

`fct_recode()` allows you to manually change the name of each level.

***#Fill in the blank***

```{r}
# Change the names of the levels to something shorter
fct_recode(responses,
  "A" = "Agree",
  "D" = "Disagree",
  "N" = "Neutral"
)
```

\newpage

#### **`fct_collapse()`: Collapse Multiple Levels into One**

`fct_collapse()` is useful for grouping several old levels into a single new level.

***#Fill in the blank***

```{r}
# Group "Agree" and "Disagree" into a new level called "Opinion"
fct_collapse(responses,
  Opinion = c("Agree", "Disagree"),
  No_Opinion = "Neutral"
)
```

\newpage

#### **`fct_lump_n()`: Lump Least Common Levels**

`fct_lump_*` functions are for collapsing the least/most common levels into a new "Other" level. `fct_lump_n()` keeps the `n` most frequent levels and lumps the rest.

***#Fill in the blank***

```{r}
# Create a more complex factor
complex_factor <- factor(c("A", "B", "A", "C", "A", "B", "D", "E"))

# Keep the 2 most frequent levels ("A" and "B"), and lump the rest into "Other"
fct_lump_n(complex_factor, n = 2)
```

------------------------------------------------------------------------

## \[END\] Video 6

\newpage

\newpage




### **Video 7) `ggplot2` - 2: Advanced Visualization with Data Transformation**

We'll start with messy, wide-format data, use tidyr and dplyr to clean and combine it, and then use advanced ggplot2 and forcats functions to create a powerful, multi-faceted visualization. We will then explore several other enhancement functions using the mtcars dataset.#Fill in the blank# Load all necessary packages

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
```

\newpage

Student Quiz Data

Step 1: The "Raw" Data Imagine we have two separate datasets. One has student quiz scores in a "wide" format, and another has student demographic information.Setup: Create the two data frames.

```{r}
student_scores_wide <- tribble(
  ~name, ~Quiz_1, ~Quiz_2,
  "Ana",       85,      90,
  "Ben",       92,      88,
  "Charlie",   78,      85
)

student_demographics <- tribble(
  ~name,      ~major,
  "Ana",   "Statistics",
  "Ben",      "Biology",
  "Charlie",  "Statistics"
)
```

Step 2: Transformation with `pivot_longer` and `left_join`

To plot this data effectively, we first need to make the scores data "long" using `pivot_longer`. Then, we'll add the major information using a left_join.

#Fill in the blank

```{r}
# First, pivot the scores data to a long format
scores_long <- student_scores_wide %>%
  pivot_longer(
    cols = c(Quiz_1, Quiz_2),
    names_to = "quiz",
    values_to = "score"
  )
# Now, join the demographic data to the long scores data
full_student_data <- left_join(scores_long, student_demographics, by = "name")
```

\newpage

Step 3: Visualization with `forcats` and ggplot2

Now that our data is tidy, we can create an advanced plot. Let's make a bar chart showing the average score for each major, ordered from highest to lowest.

`fct_reorder()`: Making Plots Better

`fct_reorder()` allows us to reorder the levels of a factor (like major) based on another variable (like the average score). This makes our plots much easier to interpret.

```{r}
# First, calculate the average score per major
avg_scores_by_major <- full_student_data %>%
  group_by(major) %>%
  summarize(avg_score = mean(score))
# Create the bar plot, using fct_reorder to sort the y-axis
ggplot(avg_scores_by_major, aes(x = avg_score, y = fct_reorder(major, avg_score))) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average Quiz Scores by Major",
    x = "Average Score",
    y = "Major"
  )
```

\newpage

`facet_wrap()`: Creating Subplots

Now let's go back to our `full_student_data` and create a more detailed visualization: a bar chart for each student, faceted by their major.

```{r}
# Create a separate plot for each major
ggplot(full_student_data, aes(x = quiz, y = score, fill = name)) +
  geom_col(position = "dodge") +
  facet_wrap(~ major) +
  labs(
    title = "Quiz Scores by Student and Major",
    x = "Quiz",
    y = "Score"
  )
```

\newpage

Techniques Showcase: Additional ggplot2 Enhancements with mtcars

Now we'll explore some other useful ggplot2 functions using the mtcars dataset.

Setup: Create a base plot for the mtcars examples.

```{r}
# We'll use this base plot for the following examples
base_plot <- ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point()
```

`coord_flip()`: Flipping Axes

`coord_flip()` swaps the x and y axes. This can be very useful for making horizontal boxplots or bar charts, which are often easier to read if you have long category labels.

```{r}
# Create a boxplot and then flip the coordinates
ggplot(mtcars, aes(x = factor(cyl), y = hp)) +
  geom_boxplot() +
  coord_flip()
```

\newpage

`scale_color_manual()`: Customizing Colors

This function gives you manual control over the colors used in your plot. You can specify the exact color for each level of your aesthetic mapping.

```{r}
# Start with our base plot and add manual colors
base_plot +
  scale_color_manual(values = c("4" = "blue", "6" = "orange", "8" = "red"))
```

\newpage

`theme()`: Customizing Plot Appearance

The `theme()` function is incredibly powerful and lets you customize almost any non-data element of your plot. Here, we'll use it to move the legend to the bottom of the plot.

```{r}
# Start with our base plot and move the legend
base_plot +
  theme(legend.position = "bottom")
```

\newpage

`geom_text()`: Adding Text Labels

This geom adds text annotations directly onto the plot. It requires `x`, `y`, and `label` aesthetics. It's often used with a filtered dataset to avoid cluttering the plot.

```{r}
# Create a subset of data for labeling
label_data <- mtcars %>%
  rownames_to_column("car") %>%
  filter(hp > 250)
```

```{r}
# Add car names to the points for very high horsepower cars
base_plot +
  geom_text(data = label_data, aes(label = car), vjust = -1)
```

\newpage

`ggsave()`: Saving Your Plot

After you've created the perfect plot, `ggsave()` allows you to save it to a file, like a PNG or PDF, with control over the size and resolution.

```{r}
#Create the final plot object
final_plot <- base_plot +
  facet_wrap(~ factor(cyl)) +
  labs(title = "MPG vs. Weight by Cylinder")
# Save the plot to a file named "my_car_plot.png"
ggsave("my_car_plot.png", plot = final_plot, width = 8, height = 5)
```

## \[END\] Video 7

\newpage

\newpage

### **Video 8) `lubridate`: Working with Dates and Times**

#### **Video Notes**

Working with dates and times can be tricky. The `lubridate` package makes it much simpler by providing functions to parse, manipulate, and extract information from date-time objects. We will create a sample data frame to demonstrate its most useful functions.

***#Fill in the blank***

```{r}
# Load the lubridate package
library(lubridate)
library(dplyr)

# Create a data frame with dates and times stored as character strings
event_log <- data.frame(
  event_name = c("Project Start", "Midterm Review", "Final Deadline"),
  start_date = c("2025-01-20", "2025-03-10", "2025-05-02"),
  start_time = c("01/20/2025 09:00:00", "03/10/2025 14:30:00", "05/02/2025 17:00:00")
)
```

\newpage

#### **1. `ymd()`: Parsing Dates**

Lubridate's parsing functions like `ymd()`, `mdy()`, `dmy()`, etc., automatically figure out and convert character strings into proper date objects. The function name corresponds to the order of the year, month, and day in your string.

***#Fill in the blank***

```{r}
# Parse the 'start_date' column, which is in Year-Month-Day format
event_log %>%
  mutate(date_parsed = ymd(start_date))
```

\newpage

#### **2. `mdy_hms()`: Parsing Datetimes**

Similarly, functions like `mdy_hms()` parse strings that contain both date and time components.

***#Fill in the blank***

```{r}
# Parse the 'start_time' column, which is in Month-Day-Year Hour:Minute:Second format
event_log %>%
  mutate(datetime_parsed = mdy_hms(start_time))
```

\newpage

#### **3 & 4. `wday()` & `month()`: Extracting Components**

Once you have a date-time object, you can easily pull out individual components like the day of the week, month, year, or hour.

***#Fill in the blank***

```{r}
# Create a parsed date column first
parsed_log <- event_log %>% mutate(date_parsed = ymd(start_date))

# Extract the day of the week and the month
parsed_log %>%
  mutate(
    day_of_week = wday(date_parsed, label = TRUE),
    event_month = month(date_parsed, label = TRUE)
  )
```

\newpage

#### **5. `duration()`: Measuring Time Spans**

Duration objects represent an exact number of seconds. You can create them with functions like `ddays()`, `dweeks()`, or by subtracting two date-time objects.

***#Fill in the blank***

```{r}
# Create two date objects
date1 <- ymd("2025-07-10")
date2 <- ymd("2025-08-25")

# Find the duration between the two dates
time_span <- duration(date2 - date1)

# Create a duration of 3 days
three_days <- ddays(3)
```

\newpage

#### **6. `now()`: Getting the Current Time**

The `now()` function returns the current system date and time.

***#Fill in the blank***

```{r}
# Get the current date and time
current_moment <- now()
print(current_moment)
```

------------------------------------------------------------------------

## \[END\] Video 8

\newpage

\newpage
