---
title: "Video 7"
format: html
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```



### **Video 7) `ggplot2` - 2: Advanced Visualization with Data Transformation**


We'll start with messy, wide-format data, use tidyr and dplyr to clean and combine it, and then use advanced ggplot2 and forcats functions to create a powerful, multi-faceted visualization. We will then explore several other enhancement functions using the mtcars dataset.#Fill in the blank# Load all necessary packages

```{r}
library(____)
library(____)
library(____)
library(____)
```

::: {style="height: 300px; background-color: white;"}
:::



Student Quiz Data

Step 1: The "Raw" Data Imagine we have two separate datasets. One has student quiz scores in a "wide" format, and another has student demographic information.Setup: Create the two data frames.

```{r}
student_scores_wide <- tribble(
  ~name, ~Quiz_1, ~Quiz_2,
  "Ana",       85,      90,
  "Ben",       92,      88,
  "Charlie",   78,      85
)

student_demographics <- tribble(
  ~name,      ~major,
  "Ana",   "Statistics",
  "Ben",      "Biology",
  "Charlie",  "Statistics"
)
```

Step 2: Transformation with `pivot_longer` and `left_join`


To plot this data effectively, we first need to make the scores data "long" using `pivot_longer`. Then, we'll add the major information using a left_join.


```{r}
# First, pivot the scores data to a long format
scores_long <- student_scores_wide %>%
  pivot_longer(
    cols = c(____, ____),
    names_to = "quiz",
    values_to = "score"
  )
```



```{r}
# Now, join the demographic data to the long scores data
full_student_data <- left_join(____, ____, by = "name")
```

::: {style="height: 300px; background-color: white;"}
:::

Step 3: Visualization with `forcats` and ggplot2

Now that our data is tidy, we can create an advanced plot. Let's make a bar chart showing the average score for each major, ordered from highest to lowest. 

`fct_reorder()`: Making Plots Better

`fct_reorder()` allows us to reorder the levels of a factor (like major) based on another variable (like the average score). This makes our plots much easier to interpret.



```{r}
# First, calculate the average score per major
avg_scores_by_major <- full_student_data %>%
  group_by(____) %>%
  summarize(avg_score = mean(score))
```

```{r}
# Create the bar plot, using fct_reorder to sort the y-axis
ggplot(avg_scores_by_major, aes(x = avg_score, y = fct_reorder(major, ____))) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average Quiz Scores by Major",
    x = "Average Score",
    y = "Major"
  )
``` 
  


::: {style="height: 300px; background-color: white;"}
:::

`facet_wrap()`: Creating Subplots

Now let's go back to our `full_student_data` and create a more detailed visualization: a bar chart for each student, faceted by their major.





```{r}
# Create a separate plot for each major
ggplot(full_student_data, aes(x = quiz, y = score, fill = name)) +
  geom_col(position = "dodge") +
  facet_wrap(~ ____) +
  labs(
    title = "Quiz Scores by Student and Major",
    x = "Quiz",
    y = "Score"
  )
``` 
  
  
::: {style="height: 300px; background-color: white;"}
:::







Techniques Showcase: Additional ggplot2 Enhancements with mtcars

Now we'll explore some other useful ggplot2 functions using the mtcars dataset.

Setup: Create a base plot for the mtcars examples.


```{r}
# We'll use this base plot for the following examples
base_plot <- ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point()
```

`coord_flip()`: Flipping Axes

`coord_flip()` swaps the x and y axes. This can be very useful for making horizontal boxplots or bar charts, which are often easier to read if you have long category labels.



```{r}
# Create a boxplot and then flip the coordinates
ggplot(mtcars, aes(x = factor(cyl), y = hp)) +
  geom_boxplot() +
  ____()
```  
  
  
::: {style="height: 300px; background-color: white;"}
:::

`scale_color_manual()`: Customizing Colors

This function gives you manual control over the colors used in your plot. You can specify the exact color for each level of your aesthetic mapping.

```{r}
# Start with our base plot and add manual colors
base_plot +
  scale_color_manual(values = c("4" = ____, "6" = ____, "8" = ____))
```
  
  
::: {style="height: 300px; background-color: white;"}
:::

`theme()`: Customizing Plot Appearance

The `theme()` function is incredibly powerful and lets you customize almost any non-data element of your plot. Here, we'll use it to move the legend to the bottom of the plot.


```{r}
# Start with our base plot and move the legend
base_plot +
  theme(legend.position = ____)
```
  
::: {style="height: 300px; background-color: white;"}
:::


`geom_text()`: Adding Text Labels

This geom adds text annotations directly onto the plot. It requires `x`, `y`, and `label` aesthetics. It's often used with a filtered dataset to avoid cluttering the plot.



```{r}
# Create a subset of data for labeling
label_data <- mtcars %>%
  rownames_to_column("car") %>%
  filter(hp > 250)
```


```{r}
# Add car names to the points for very high horsepower cars
base_plot +
  geom_text(data = ____, aes(label = car), vjust = -1)
```

::: {style="height: 300px; background-color: white;"}
:::

`ggsave()`: Saving Your Plot

After you've created the perfect plot, `ggsave()` allows you to save it to a file, like a PNG or PDF, with control over the size and resolution.


```{r}
#Create the final plot object
final_plot <- base_plot +
  facet_wrap(~ factor(cyl)) +
  labs(title = "MPG vs. Weight by Cylinder")
```

```{r}
# Save the plot to a file named "my_car_plot.png"
ggsave(____, plot = ____, width = 8, height = 5)
```



## \[END\] Video 7

::: {style="height: 300px; background-color: white;"}
:::

::: {style="height: 300px; background-color: white;"}
:::
