---
title: "Video 8"
format: html
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


### **Video 8) `lubridate`: Working with Dates and Times**

#### **Video Notes**

Working with dates and times can be tricky. The `lubridate` package makes it much simpler by providing functions to parse, manipulate, and extract information from date-time objects. We will create a sample data frame to demonstrate its most useful functions.

***#Fill in the blank***

```{r}
## Load the lubridate package
# library(____)
# library(dplyr)
# 
## Create a data frame with dates and times stored as character strings
# event_log <- data.frame(
#   event_name = c("Project Start", "Midterm Review", "Final Deadline"),
#   start_date = c("2025-01-20", "2025-03-10", "2025-05-02"),
#   start_time = c("01/20/2025 09:00:00", "03/10/2025 14:30:00", "05/02/2025 17:00:00")
# )
```

::: {style="height: 300px; background-color: white;"}
:::

#### **1. `ymd()`: Parsing Dates**

Lubridate's parsing functions like `ymd()`, `mdy()`, `dmy()`, etc., automatically figure out and convert character strings into proper date objects. The function name corresponds to the order of the year, month, and day in your string.

***#Fill in the blank***

```{r}
## Parse the 'start_date' column, which is in Year-Month-Day format
# event_log %>%
#   mutate(date_parsed = ymd(____))
```

::: {style="height: 300px; background-color: white;"}
:::

#### **2. `mdy_hms()`: Parsing Datetimes**

Similarly, functions like `mdy_hms()` parse strings that contain both date and time components.

***#Fill in the blank***

```{r}
## Parse the 'start_time' column, which is in Month-Day-Year Hour:Minute:Second format
# event_log %>%
#   mutate(datetime_parsed = mdy_hms(____))
```

::: {style="height: 300px; background-color: white;"}
:::

#### **3 & 4. `wday()` & `month()`: Extracting Components**

Once you have a date-time object, you can easily pull out individual components like the day of the week, month, year, or hour.

***#Fill in the blank***

```{r}
## Create a parsed date column first
# parsed_log <- event_log %>% mutate(date_parsed = ymd(start_date))
# 
## Extract the day of the week and the month
# parsed_log %>%
#   mutate(
#     day_of_week = wday(____, label = TRUE),
#     event_month = month(____, label = TRUE)
#   )
```

::: {style="height: 300px; background-color: white;"}
:::

#### **5. `duration()`: Measuring Time Spans**

Duration objects represent an exact number of seconds. You can create them with functions like `ddays()`, `dweeks()`, or by subtracting two date-time objects.

***#Fill in the blank***

```{r}
## Create two date objects
# date1 <- ymd("2025-07-10")
# date2 <- ymd("2025-08-25")
# 
## Find the duration between the two dates
# time_span <- duration(____ - ____)
# 
## Create a duration of 3 days
# three_days <- ddays(____)
```

::: {style="height: 300px; background-color: white;"}
:::

#### **6. `now()`: Getting the Current Time**

The `now()` function returns the current system date and time.

***#Fill in the blank***

```{r}
## Get the current date and time
# current_moment <- now()
# print(current_moment)
```

------------------------------------------------------------------------

## \[END\] Video 8

::: {style="height: 300px; background-color: white;"}
:::

::: {style="height: 300px; background-color: white;"}
:::
