---
title: "Video 5"
format: html
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


### **Video 5) `stringr`: Working with Text Data**

#### **Video Notes**

In this video, we'll explore the `stringr` package, which provides a consistent and easy-to-use set of tools for working with character strings. We'll load the `stringr` package and create a sample vector of text to work with for our examples.

***#Fill in the blank***

```{r}
# Load the stringr package
# library(____)
# 
# # Create a vector of sample sentences
# sentences <- c(
#   "The quick brown fox jumps over the lazy dog.",
#   "R is a language for statistical computing.",
#   "The year is 2025, and the sky is blue.",
#   "Another sentence with the word dog."
# )
```

::: {style="height: 300px; background-color: white;"}
:::

#### **`str_detect()`: Detect a Pattern**

`str_detect()` returns a logical (TRUE/FALSE) vector indicating whether a pattern is found in each string.

***#Fill in the blank***

```{r}
## Detect which sentences contain the word "dog"
# str_detect(____, pattern = ____)
```

::: {style="height: 300px; background-color: white;"}
:::

#### **`str_replace_all()`: Replace Patterns**

`str_replace_all()` finds all matches of a pattern and replaces them with a new string.

***#Fill in the blank***

```{r}
## Replace all instances of the word "dog" with "cat"
# str_replace_all(____, pattern = ____, replacement = ____)
```

::: {style="height: 300px; background-color: white;"}
:::

#### **`str_to_upper()`: Change Case**

Functions like `str_to_upper()`, `str_to_lower()`, and `str_to_title()` change the case of your strings.

***#Fill in the blank***

```{r}
## Convert the first sentence to all uppercase
# str_to_upper(____)
```

::: {style="height: 300px; background-color: white;"}
:::

#### **`str_subset()`: Keep Strings with a Pattern**

`str_subset()` is a useful shortcut that returns only the strings from your vector that contain a match for the pattern.

***#Fill in the blank***

```{r}
## Keep only the sentences that contain the word "dog"
# str_subset(____, pattern = ____)
```

::: {style="height: 300px; background-color: white;"}
:::

#### **`str_extract()`: Extract a Pattern**

`str_extract()` pulls out the first occurrence of a pattern. This is powerful when combined with regular expressions to find specific types of text, like numbers. The pattern `\\d+` matches one or more digits.

***#Fill in the blank***

```{r}
## Extract the number from our sentences vector
# str_extract(____, pattern = "\\d+")
```

::: {style="height: 300px; background-color: white;"}
:::

#### **`str_split()`: Split a String**

`str_split()` breaks a string into a list of smaller pieces based on a delimiter pattern.

***#Fill in the blank***

```{r}
## Split the first sentence into a vector of individual words
## The pattern " " splits the string at each space
# str_split(____, pattern = ____)
```

------------------------------------------------------------------------

## \[END\] Video 5

::: {style="height: 300px; background-color: white;"}
:::

::: {style="height: 300px; background-color: white;"}
:::
