---
title: "Video 6"
format: html
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


### **Video 6) `forcats`: Working with Factors**

#### **Video Notes**

This video covers the `forcats` package, which is designed for working with factors. Factors are R's way of handling categorical variables, and `forcats` provides tools to solve common problems, like changing the order of levels or re-grouping them.

***#Fill in the blank***

```{r}
## Load the forcats package
# library(____)

## Create a sample factor representing survey responses
# responses <- factor(c("Agree", "Disagree", "Agree", "Neutral", "Disagree", "Agree"))
```

::: {style="height: 300px; background-color: white;"}
:::

#### **`fct_relevel()`: Manually Reorder Levels**

By default, factor levels are ordered alphabetically. `fct_relevel()` lets you move specific levels to the front, changing their order for plots and tables.

***#Fill in the blank***

```{r}
##The original order is alphabetical: "Agree", "Disagree", "Neutral"
# levels(responses)

## Move "Neutral" to be the second level, after "Agree"
# fct_relevel(____, "Neutral", after = ____)
```

::: {style="height: 300px; background-color: white;"}
:::

#### **`fct_infreq()`: Reorder by Frequency**

`fct_infreq()` is a useful shortcut to reorder factor levels so that the most common level appears first.

***#Fill in the blank***

```{r}
## Reorder the levels based on how many times each one appears
# fct_infreq(____)
```

::: {style="height: 300px; background-color: white;"}
:::

#### **`fct_rev()`: Reverse Level Order**

`fct_rev()` simply reverses the current order of the levels.

***#Fill in the blank***

```{r}
## Reverse the order of the levels in our responses factor
# fct_rev(____)
```

::: {style="height: 300px; background-color: white;"}
:::

#### **`fct_recode()`: Change Level Values**

`fct_recode()` allows you to manually change the name of each level.

***#Fill in the blank***

```{r}
## Change the names of the levels to something shorter
# fct_recode(____,
#   "A" = ____,
#   "D" = ____,
#   "N" = ____
# )
```

::: {style="height: 300px; background-color: white;"}
:::

#### **`fct_collapse()`: Collapse Multiple Levels into One**

`fct_collapse()` is useful for grouping several old levels into a single new level.

***#Fill in the blank***

```{r}
## Group "Agree" and "Disagree" into a new level called "Opinion"
# fct_collapse(____,
#   Opinion = c(____, ____),
#   No_Opinion = ____
# )
```

::: {style="height: 300px; background-color: white;"}
:::

#### **`fct_lump_n()`: Lump Least Common Levels**

`fct_lump_*` functions are for collapsing the least/most common levels into a new "Other" level. `fct_lump_n()` keeps the `n` most frequent levels and lumps the rest.

***#Fill in the blank***

```{r}
# Create a more complex factor
complex_factor <- factor(c("A", "B", "A", "C", "A", "B", "D", "E"))

# Keep the 2 most frequent levels ("A" and "B"), and lump the rest into "Other"
# fct_lump_n(____, n = ____)
```

------------------------------------------------------------------------

## \[END\] Video 6

::: {style="height: 300px; background-color: white;"}
:::

::: {style="height: 300px; background-color: white;"}
:::