---
title: "ETV Cycle 5 - NBA GOATs: Excel Sheets, Joins, and Advanced ggplot2 (EMPTY)"
format: docx
editor: visual
---

## Description

This is the official Cycle 5 ETV workflow. We return to the NBA GOATs Excel workbook and use the full **Extract -\> Transform -\> Visualize** structure in one cohesive analysis.

> **Note:** Extraction looks different in this cycle — we use `map()` to read all sheets in a single step. The emphasis this cycle falls on **Visualization** — the plots in Part V directly answer the motivating questions using division and conference information attached during the Transformation step.

The workbook has one sheet per player, plus two support sheets:

-   `KB`, `LJ`, and `MJ`: game-level data for Kobe Bryant, LeBron James, and Michael Jordan.
-   `variable_names`: a data dictionary describing the workbook columns.
-   `divisions`: a lookup table connecting team abbreviations to team, division, and conference.

Each **row** in the combined player data represents one game. Each **column** represents a game-level statistic or game descriptor.

------------------------------------------------------------------------

### Motivating Questions

**Subtask 1** - Which player-team pairs produced the highest average Game Score, and does conference membership (East vs. West) affect that ranking?

**Subtask 2** - How do we attach division and conference information to each player's game record, and how can we verify there are no mismatches before joining?

**Subtask 3** - Does a player's scoring distribution look different when facing Eastern Conference opponents compared to Western Conference opponents?

**Subtask 4** - Are home-game win rates different from away-game win rates, and does that pattern hold the same way for Kobe Bryant, LeBron James, and Michael Jordan?

------------------------------------------------------------------------

## Learning Moment 1 - Libraries

```{r}
#| message: false
#| warning: false
library(here)
library(readxl)
library(purrr)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(forcats)
library(scales)
```

------------------------------------------------------------------------

## Extraction (E)

#### Learning Moment 2 - Inspecting Sheet Names

```{r}
path_to_goats <- here("ETV", "Data", "nba_goats.xlsx")

sheet_names <- excel_sheets(path_to_goats)
sheet_names
```

------------------------------------------------------------------------

#### Learning Moment 3 - Reading All Sheets with `map()`

`all_sheets` is a named **list** — each element is one data frame and the names come from the sheet names.

```{r}
all_sheets <- sheet_names |>
  set_names() |>
  map(\(sheet) read_excel(path_to_goats, sheet = sheet))

names(all_sheets)
```

```{r}
all_sheets[["variable_names"]] |>
  slice_head(n = 8)
```

------------------------------------------------------------------------

## Transformation (T)

#### Learning Moment 4 - Binding Player Sheets Before Joining

```{r}
player_sheets <- c("KB", "LJ", "MJ")

read_player_sheet <- function(sheet_name) {
  read_excel(path_to_goats, sheet = sheet_name) |>
    mutate(
      source_sheet = sheet_name,
      player = recode(
        Name,
        KB = "Kobe Bryant",
        LJ = "LeBron James",
        MJ = "Michael Jordan"
      ),
      .before = 1
    )
}

nba_players_df <- player_sheets |>
  map_dfr(read_player_sheet)

glimpse(nba_players_df)
```

**Q3.** Why do we bind the three player sheets before joining to division data?

> Your answer:

------------------------------------------------------------------------

#### Learning Moment 5 - Why We Need a Join Here

In this workflow, player game data and division/conference information live in separate sheets. We need a join to bring them together.

**Why is the join needed?**

-   Each game row records the team abbreviation (`Tm`) but not the division or conference.
-   The `divisions` sheet is a lookup table — one row per team, with the full team name, division, and conference.
-   A `left_join` adds division and conference to every player-game row by matching on the team abbreviation.
-   We use `left_join` specifically because we want to keep **every** game row in the player data, even if a team abbreviation is unusual or missing from the lookup table.

This is the only join in the workflow. Before running it, we verify that all team abbreviations match using `anti_join()` (see Learning Moment 6).

------------------------------------------------------------------------

#### Learning Moment 6 - Preparing the Division Lookup Table

```{r}
division_df <- all_sheets[["divisions"]] |>
  rename(
    team = Abbreviation,
    team_name = `Full Team Name`,
    division = Division,
    conference = Conference
  )

division_df |>
  slice_head(n = 8)
```

```{r}
nba_players_df |>
  distinct(Tm) |>
  anti_join(division_df, by = c("Tm" = "team"))
```

**Q5.** Why do we run this `anti_join()` before the real join?

> Your answer:

------------------------------------------------------------------------

#### Learning Moment 7 - Joining Division Data to Player Data

```{r}
nba_joined_df <- nba_players_df |>
  left_join(division_df, by = c("Tm" = "team"))

nba_joined_df |>
  select(player, Season, Tm, team_name, division, conference, PTS, AST, TRB, Game_Outcome) |>
  slice_head(n = 10)
```

```{r}
nba_joined_df |>
  count(player, team_name, division, conference, sort = TRUE)
```

**Q6.** What does `by = c("Tm" = "team")` tell `left_join()`?

> Your answer:

------------------------------------------------------------------------

#### Learning Moment 8 - Creating Analysis Variables

```{r}
nba_analysis_df <- nba_joined_df |>
  mutate(
    season_num = as.integer(str_extract(Season, "\\d+")),
    outcome = if_else(Game_Outcome == "W", "Win", "Loss"),
    home_away = Game_Location,
    points_group = case_when(
      PTS >= 40 ~ "40+ points",
      PTS >= 30 ~ "30-39 points",
      PTS >= 20 ~ "20-29 points",
      TRUE ~ "Under 20 points"
    ),
    points_group = factor(
      points_group,
      levels = c("Under 20 points", "20-29 points", "30-39 points", "40+ points")
    )
  )

nba_analysis_df |>
  select(player, season_num, team_name, division, conference, outcome, PTS, AST, TRB, GmSc, points_group) |>
  slice_head(n = 8)
```

**Q7.** Why is it useful to create `season_num`, `outcome`, and `points_group` before plotting?

> Your answer:

------------------------------------------------------------------------

## Visualization (V)

Prior cycles already used horizontal bar charts, error bars, histograms, log scales, scatter plots, line plots, faceting, and manual color scales. In this cycle, focus on labels, annotations, percent scales, themes, and faceted comparisons. Each subtask directly answers one of the motivating questions.

**Style rule:** pipe the data frame into the plot: `df |> ggplot(aes(...))`.

------------------------------------------------------------------------

#### Subtask 1 - Which player-team pair produced the highest average Game Score, and does conference membership affect that ranking?

Build the plot in two steps: run the base layer first, then add layers to produce the final answer.

```{r}
player_team_gmsc_df <- nba_analysis_df |>
  group_by(player, team_name, conference) |>
  summarize(
    games = n(),
    avg_game_score = mean(GmSc, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(games >= 50)

# Step 1 — Base bar chart
p1_base <- player_team_gmsc_df |>
  ggplot(aes(
    x = fct_reorder(team_name, avg_game_score),
    y = avg_game_score,
    fill = conference
  )) +
  geom_col(width = 0.75)

p1_base
```

```{r}
# Step 2 — Add layers to reach the target plot
# Hint: add geom_text() for value labels, coord_flip(), scale_y_continuous() limits,
#        labs() for titles and axis labels, and theme() for polish
p1_base +
  ## Your code here

```

**Q8.** What do `geom_text()` and `scale_y_continuous()` add to this plot?

> Your answer:

------------------------------------------------------------------------

#### Subtask 2 - Does scoring distribution look different when facing Eastern vs. Western Conference opponents?

```{r}
nba_analysis_df |>
  filter(!is.na(PTS), !is.na(conference)) |>
  ggplot(aes(x = conference, y = PTS, fill = conference)) +
  geom_boxplot(alpha = 0.75, outlier.alpha = 0.25) +
  facet_wrap(~ player) +
  scale_y_continuous(breaks = breaks_width(10)) +
  labs(
    title = "Scoring Distribution by Conference",
    subtitle = "Each panel shows one player",
    x = NULL,
    y = "Points Scored",
    fill = "Conference"
  ) +
  theme_light(base_size = 12) +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "gray90", color = NA),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

**Q9.** What does a boxplot show that a bar chart of averages would hide?

> Your answer:

------------------------------------------------------------------------

#### Subtask 3 - Are scoring and assists related differently across divisions?

```{r}
nba_analysis_df |>
  filter(!is.na(PTS), !is.na(AST), !is.na(division)) |>
  ggplot(aes(x = AST, y = PTS, color = division)) +
  geom_point(alpha = 0.35, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  facet_wrap(~ player) +
  annotate(
    "text",
    x = 14,
    y = 55,
    label = "High assist games can still include elite scoring",
    hjust = 1,
    size = 3
  ) +
  scale_x_continuous(breaks = breaks_width(4)) +
  scale_y_continuous(breaks = breaks_width(10)) +
  labs(
    title = "Points and Assists by Player",
    subtitle = "Color identifies the division attached to the player's team",
    x = "Assists",
    y = "Points",
    color = "Division"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

**Q10.** Why might `alpha = 0.35` help in this scatter plot?

> Your answer:

------------------------------------------------------------------------

#### Subtask 4 - Are home-game win rates different from away-game win rates across players?

Build the plot in steps: start with a base layer, then add percentage labels and per-player team colors.

```{r}
# Step 1 — Compute win/loss proportions, then build a per-player fill key
# Each player gets their own Win/Loss palette, so we combine player and
# outcome into a single fill variable (e.g. "Michael Jordan_Win").
outcome_location_df <- nba_analysis_df |>
  filter(!is.na(outcome), !is.na(home_away)) |>
  count(player, home_away, outcome) |>
  group_by(player, home_away) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  mutate(fill_key = paste(player, outcome, sep = "_"))

# Step 2 — Base stacked bar chart (fill on the combined key)
p4_base <- outcome_location_df |>
  ggplot(aes(x = home_away, y = prop, fill = fill_key)) +
  geom_col(position = "stack", color = "white", width = 0.7)

p4_base
```

```{r}
# Step 3 — Add layers to reach the target plot
# Hint: add geom_text() with percent() labels (position_stack vjust = 0.5,
#        color mapped to outcome for readable contrast),
#        facet_wrap(), scale_fill_manual() with per-player team colors
#        (Jordan: black/#CE1141, LeBron: #860038/#FDBB30, Kobe: #552583/#FDB927),
#        scale_color_manual() for text contrast, scale_y_continuous(),
#        labs(), and theme()
p4_base +
  ## Your code here

```

**Q11.** Why is a percent scale useful here?

> Your answer:

------------------------------------------------------------------------

### Student-Generated Extension

Create one new analytical question that uses `nba_analysis_df` and at least one visualization feature emphasized in this cycle:

-   scale modification,
-   theme customization,
-   labels or annotations,
-   faceting,
-   color mapping,
-   or another advanced `ggplot2` element.

```{r}
## Your code here

```

**Reflection.** What conclusion does your plot support, and what visual evidence supports that conclusion?

> Your answer:
