---
title: "ETV Cycle 5 - NBA GOATs Shiny App"
format: docx
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

## Goal

In the main Cycle 5 lesson, we built a joined data frame by reading multiple Excel sheets, binding the player sheets, and joining division information. In this extension, we turn that workflow into a Shiny app.

The app is designed to answer several kinds of analytical questions:

- **Comparison questions:** Which player or team has the highest average points, assists, rebounds, or Game Score?
- **Distribution questions:** How spread out are player game statistics?
- **Relationship questions:** Are points and assists related, and does the pattern differ by player or conference?
- **Filtering questions:** How do conclusions change when we focus on selected players, conferences, teams, or game outcomes?

The finished app is saved in this folder as `app.R`.

------------------------------------------------------------------------

## Step 1 - Load Libraries

```{r}
library(shiny)
library(readxl)
library(purrr)
library(dplyr)
library(stringr)
library(ggplot2)
library(forcats)
library(scales)
```

------------------------------------------------------------------------

## Step 2 - Create a Data-Loading Function

The app should build the same joined data set every time it opens. We use a function so that the extraction and transformation workflow is contained in one place.

```{r}
find_workbook <- function() {
  candidates <- c(
    file.path("ETV", "Data", "nba_goats.xlsx"),
    file.path("..", "Data", "nba_goats.xlsx"),
    file.path("..", "..", "ETV", "Data", "nba_goats.xlsx")
  )

  existing <- candidates[file.exists(candidates)]

  if (length(existing) == 0) {
    stop("Could not find ETV/Data/nba_goats.xlsx from the app working directory.")
  }

  normalizePath(existing[[1]])
}
```

```{r}
load_nba_data <- function() {
  path_to_goats <- find_workbook()
  player_sheets <- c("KB", "LJ", "MJ")

  division_df <- read_excel(path_to_goats, sheet = "divisions") |>
    rename(
      team = Abbreviation,
      team_name = `Full Team Name`,
      division = Division,
      conference = Conference
    )

  player_df <- player_sheets |>
    map_dfr(\(sheet_name) {
      read_excel(path_to_goats, sheet = sheet_name) |>
        mutate(
          source_sheet = sheet_name,
          player = recode(
            Name,
            KB = "Kobe Bryant",
            LJ = "LeBron James",
            MJ = "Michael Jordan"
          ),
          .before = 1
        )
    })

  player_df |>
    left_join(division_df, by = c("Tm" = "team")) |>
    mutate(
      season_num = as.integer(str_extract(Season, "\\d+")),
      outcome = if_else(Game_Outcome == "W", "Win", "Loss"),
      home_away = Game_Location
    )
}
```

**Q1.** Why is the data-loading workflow wrapped in a function?

> Your answer:

------------------------------------------------------------------------

## Step 3 - Plan the Inputs

The app uses widgets to let the user change the analytical question without editing code.

```{r}
checkboxGroupInput("players", "Players", choices = player_choices)
selectInput("conference", "Conference", choices = conference_choices)
selectInput("stat", "Statistic", choices = stat_choices)
selectInput("group_var", "Compare by", choices = group_choices)
sliderInput("min_games", "Minimum games per group", min = 1, max = 400, value = 25)
```

**Q2.** Which input would you change to compare divisions instead of conferences?

> Your answer:

------------------------------------------------------------------------

## Step 4 - Plan the Outputs

The app has four main outputs:

- a summary bar chart,
- a distribution plot,
- a relationship plot,
- and a filtered data table.

In Shiny, each output has two parts:

```{r}
plotOutput("summary_plot")

output$summary_plot <- renderPlot({
  # code that creates the plot
})
```

**Q3.** Why must the output id `"summary_plot"` match in both places?

> Your answer:

------------------------------------------------------------------------

## Step 5 - Use a Reactive Filter

A reactive expression stores a result that updates when inputs change. We filter once, then reuse the filtered data in several outputs.

```{r}
filtered_data <- reactive({
  req(input$players)

  nba_analysis_df |>
    filter(
      player %in% input$players,
      input$conference == "All" | conference == input$conference,
      input$outcome == "All" | outcome == input$outcome
    )
})
```

**Q4.** Why is `filtered_data()` written with parentheses inside `renderPlot()`?

> Your answer:

------------------------------------------------------------------------

## Step 6 - Run the Finished App

The finished app is in `ETV/ETV-Cycle-5/app.R`.

From the project root, run:

```{r}
shiny::runApp("ETV/ETV-Cycle-5")
```

From inside the `ETV/ETV-Cycle-5` folder, run:

```{r}
shiny::runApp()
```

------------------------------------------------------------------------

## Extension Questions

Use the app to answer at least two questions.

**Question A.** Which player-team or player-division group has the highest average Game Score?

> Your answer:

**Question B.** Does the points-versus-assists relationship look similar for all three players?

> Your answer:

**Question C.** How does the answer change when filtering to wins only?

> Your answer:

**Question D.** Create your own app-based question and answer it using one plot and one table output.

> Your answer:

