---
title: "Video 1"
format: html
editor: visual
---

## **Video 1) R & Basics with `mtcars`**

#### **Video Notes - Packages & Libraries**

This section covers the fundamental concepts of R. We will not load the entire `tidyverse` package; instead, we will load individual packages as needed to better understand the function of each one.

First, let's inspect the `mtcars` dataset.

```{r}
# Look at the structure of the mtcars dataset
# str(____)
```

::: {style="height: 300px; background-color: white;"}
:::

#### **Video Notes - Look at structure Data**

The `str()` function shows us the structure of our data. Notice the `vs` (V/S engine) and `am` (transmission) variables are listed as `num` (numeric) but only contain 0s and 1s. These are **binary** variables. To learn more about them, we can use the help function by typing `?mtcars` in the console. For our purposes, we can think of these as character or factor variables (e.g., "automatic" vs. "manual"), and we'll learn how to work with them later.

To see the first few rows of the data, we use the `head()` function.

***#Fill in the blank***

```{r}
# The head() function displays the first six rows of the data set
# head(____)
```

::: {style="height: 300px; background-color: white;"}
:::

#### **Video Notes - View Data**

If you want to see the entire interactive data frame in a new tab, you can use the `View()` function (with a capital "V"). Running `View(mtcars)` in your console will open it.

------------------------------------------------------------------------

::: {style="height: 300px; background-color: white;"}
:::

#### **Data Structures**

**a) Vectors**: A vector is a one-dimensional sequence of data elements of the same type. This type can be numeric, character (string), or logical (boolean).

-   **Numeric**: Numbers (e.g., 10.5, 55, -3)
-   **Character**: Text, also known as strings (e.g., "hello", "R")
-   **Logical**: TRUE or FALSE values, also known as booleans.

***#Fill in the blank***

```{r}
# # Create a numeric vector of miles-per-gallon (mpg) values
# mpg_vector <- ____
# # Create a character vector of the car names
# car_names_vector <- ____
# # Create a boolean vector indicating if a car has more than 150 horsepower
# powerful_boolean_vector <- ____
```

::: {style="height: 300px; background-color: white;"}
:::

**b) Data Frames**: This is the primary structure for tabular data in R. `mtcars` is a classic example.

#### **Video Notes - Tibble and Data Frames**

While I primarily use data frames because that's what I'm used to, you will often see another modern data frame structure online called a **tibble**. They behave very similarly but have some friendlier printing and subsetting features.

***#Fill in the blank***

```{r}
# # Check the class of the mtcars object
# class(____)
```

::: {style="height: 300px; background-color: white;"}
:::

**c) Lists**: A list is a versatile R object that can contain elements of different types.

#### **Video Notes - Lists**

If you're familiar with programming, a **named list** in R is very similar to a **JSON** object or a **dictionary** in Python. There are many ways of extracting information from lists; here are a few common ones.

***#Fill in the blank***

```{r}
# # Create a list containing a numeric vector and a data frame
# my_list <- list(
#   description = ____,
#   mpg_stats = ____,
#   sample_cars = ____
# )
# 
# # Extract elements from the list
# # Method 1: Using the $ operator
# description <- ____
# 
# # Method 2: Using double square brackets [[ ]] with the name
# mpg_summary <- ____
# 
# # Method 3: Using double square brackets [[ ]] with the index
# first_two_cars <- ____
```

------------------------------------------------------------------------